### 2017-04-24 manuscript/draft.Rmd
results/discussion

---
output:
  pdf_document: default
  html_document: default
---
# Moving towards large-scale data-intensive forecasting of species richness in breeding birds
# Using best practices for forecasting biodiversity changes in breeding birds
Authors: Harris, Taylor and White?

`r settings = yaml::yaml.load_file("settings.yaml")`

## Introduction

Forecasting the future state of ecological systems is increasingly important for
planning, management, and evaluating how well ecological models capture the key
ecological processes governing a system [@clark2001, @dietz2017]. One area of
particular interest for forecasting is biodiversity since it is important for
ecosystem function, central to conservation planning, and expected to be
influenced by anthropogenic global change. Anticipating potential changes in
biodiversity is crucial for large scale management and conservation, and for
addressing debates regarding whether local scale diversity is declining and may
continue to decline in response to anthropogenic influences [@dornelas2014].

Previous efforts to predict future patterns of species richness, and diversity
more generally, have focused primarily on building species distributions models
to quantify the spatial relationships between the occurrence of individual
species and the environment. Forecasts for future environmental conditions are
then used to predict where each species will occur in the future and the
probabilities of occurrence are summed across species to predict future species
richness [e.g., @thomas2004]. Alternatively, models that directly relate spatial
patterns of species richness to the environment have been developed and
generally perform equivalently to species distribution modeling based methods
[@algar2009, @distler2015].

However, despite the emerging interest in forecasting species richness and other
aspects of biodiversity, little is known about how effectively we can anticipate
their dynamics. This is due in part to the long time scales over which many
ecological forecasts are applied and the resulting difficulty in assessing
whether the predicted changes occurred. What we do know comes from a small
number of hindcasting studies, where models are built using data on species
occurrence and richness from the past and evaluated on their ability to predict
contemporary patterns [e.g., @algar2009, @distler2015]. These studies are a
valuable first step, but lack several components that are important for
developing good forecasting models and understanding how accurately these models
can predict the future. These components of good forecasting and evaluation (Box
1) broadly involve: 1) expanding the use of data to include biological and
environmental time-series [@treddenick2016]; 2) accounting for uncertainty in
observations, processes, model choice, and forecast evaluation
[@clark2001, @dietz2017]; and 3) conducting meaningful evaluations of the
forecasts by hindcasting, archiving short-term forecasts, and comparing
forecasts to baselines to determine if the forecasts are more accurate than
assuming the system is basically static [@Perretti2013].

In this paper we attempt to forecast the species richness of breeding birds at
hundreds of locations throughout North America while following the best
practices in Box 1 for ecological forecasting. To do this we combine 30 years of
time-series data on bird distributions with monthly time-series of climate data
and satellite-based remote-sensing. This 30 year time-scale has only recently
become possible for large-scale time-series based forecasting and allows us to
model and assess changes a decade or more into the future in the presence of
shifts in environmental conditions on par with predicted climate change. We
compare traditional distribution modeling based approaches to spatial models of
species richness, time-series methods, and simple baselines. All forecasts model
uncertainty and observation, are evaluated across different time lags using
hindcasting, and are publicly archived to allow future assessment. We discuss
the implications of these practices for our understanding of, and confidence in,
the resulting forecasts, and how we can continue to improve on these approaches
in the future.

## Methods

### Data

* BBS + retriever
* Predictor variables
  * Monthly climate data from interpolated weather stations (PRISM), converted
    to BIOCLIM variables. Many of these variables were tightly correlated, so
    we only included the 8 variables selected by Harris's (2015) BBS analysis
  * Weather/climate and NDVI were each calculated separately for each year,
    rather than averaged over the full multi-decade period (as Worldclim does)
  * NDVI (GIMMS)
  * Altitude (SRTM 90m Digital Elevation Database via raster::getData)
* Data summary: 
  * Number of sites, species, unique observers
  * central moments of richness (mean about 51, sd about 12 spp)
  * The within-site variation was much smaller (sd about 5 spp), making the
    site-level distributions underdispersed.  
    * This underdispersion implies
      that common count models such as binomial, Poisson, negative binomial 
      won't fit well.
    * For this reason (and because the time series methods assumed 
      Gaussian errors), we defined model performance throughout this paper in 
      terms of Gaussians (e.g. mean squared error), rather than with count
      distributions.
        * Some of the modeling approaches (SDMs and JSDMs) did use discrete
          distributions internally, however.
* Splitting the data temporally
  * Our data set began in `r settings$start_yr`, which was the first year that 
    NDVI data was available from GIMMS
  * All the models were trained on the data that was collected through 
    `r settings$last_train_year`
  * Sites were included in the analysis if there were at least 
    `r settings$min_num_yrs` years of data available during this period.
  * The models were tested/evaluated on the data from 
    `r settings$last_train_year + 1` `r settings$end_yr`.

### Accounting for site \& observer effects
* Observer effects are inherent in large data sets taken by different 
  observers, and are known to occur in BBS [citation].
* While these biases are known to exist, they are difficult to quantify: even if
  the different observers ran the same transects under identical conditions (which 
  they don't), we would still only have an estimate of their differences, not an 
  exact value.
* We estimated the observer effects, along with our uncertainty about those
  effects' size & direction, with a linear mixed model.
* This model partitions the variance in observed richness values into site-level
  variance, observer-level variance, and residual variance (e.g. variation within
  a site from year to year or from morning to morning).
  * No fixed effects in this model (e.g. environmental predictor variables because 
    we're not interested in explaining differences among sites at this point; we 
    just want to account for their existence while estimating observer effects.
  * Subtracting off the estimated observer effect provides an estimate of how many
    would have been found by a "typical" observer on that day; in this sense, it 
    enables us to correct for observer differences. We then used this "corrected" 
    version of the data for our time-series models such as the AR1, etc.
* A simple model like this one can't tell us exactly how each observer differed from
  the others, so it is important to represent our uncertainty about these differences.
  Here, we did so by collecting [[N]] Monte Carlo samples from the model's posterior
  distribution, using Stan. This provided us with [[N]] plausible values for each 
  observer, so we did not have to commit to a single value.
* While some methods for predicting richness might be able to use the uncertain
  estimates of the observer effects directly, we take the simpler approach of just
  fitting each of the downstream models with each of the Monte Carlo samples. This
  increases the amount of computational work by a factor of [[N]], but CPU time
  wasn't our limiting resource.

### Site-level models

[[note that these all use a Gaussian error distribution]]

**Baseline models.** The two simplest models we fit treated site-level richness
observations either as uncorrelated noise around a site-level constant (the 
"average" model) or as an autoregressive model with a single year of history 
(the "naive" model). The predictions from these two models are qualitatively 
different: predictions from the "average" model are centered on the average 
richness and the confidence intervals are narrow and constant-width. The
"naive" model, in contrast, predicts that future observations will be similar to
the final observed value (e.g. the value observed in `$r settings$last_train_year`), 
and the confidence intervals expand rapidly as the predictions extend farther into 
the future.  Each of these model types was fit separately to each BBS route in two
ways: first, we fit the model to the raw richness values, without accounting for
possible differences among observers at the site.  Then, we fit the models to the
residuals from our observer model (i.e. fitting the model to an estimate of how
many species a typical observer would have found, rather than to the raw data). 
In the latter case, we averaged across [[N]] models that were fit separately to 
the [[N]] Monte Carlo estimates of the observer effects, to account for our
uncertainty in the true values of those effects.

**Time series models.** Like the two baseline models, these models only looked at 
one site at a time, but they could include additional complexity. As above, these
models were trained both with the raw richness time series and with Monte Carlo
estimates of the observer-corrected time series. The  Auto-ARIMA models (based on 
Hyndman's 2016 `auto.arima` function) could include an autoregressive component 
(as in the "naive" model, but with the possibility of longer-term dependencies
in the underlying process), a moving average component (where the noise can have 
serial autocorrelation) and an integration/differencing component (so that the 
analysis could be performed on sequential differences of the raw data, rather than
on the observed time series itself). The `auto.arima` function chooses whether to 
include each of these components, and how many terms to include for each one, using
AICc, as described in Hyndman (2016). We used the default settings for this function,
apart from setting `seasonal = FALSE` because our time series does not include a 
seasonal component.  [[Note: should probably record somewhere what order model
the auto.arima function tends to choose.  Eyeballing the results, seems to be most 
similar to the naive model, but would be good to check.]] We also fit an 
"Auto-ARIMA + environment" model, in which each site-level time series could 
also be predicted by annual changes in our environmental variables at that site.

### Continental-scale models

So far, each model has been fit to one site at a time, with no information 
(apart from observer effects) shared among sites. When continental-scale data 
is available, ecologists often use correlative models to predict richness 
changes over time from environmental data. Here, we tested three variants of 
this approach. Here, we did *not* include the site-level random effects as 
predictors, meaning that this approach implicitly assumes that two sites 
with identical Bioclim, elevation, and NDVI values should have identical 
richness distributions. As above, we included observer effects & the 
associated uncertainty by running this model [[N]] times (once per MCMC 
sample).

**"Macroecological" model.** The most straightforward way to predict richness is 
to fit a richness model. We chose the `gbm` package for fitting boosted 
regression trees as our example of such a model. Boosted regression trees is a 
common approach in the species distribution modeling literature [refs] and 
works by fitting thousands of small tree-structured models sequentially, with
each tree optimized to reduce the error of its predecessors. For the reasons 
discussed above, this model was optimized using a Gaussian likelihood. [[other 
gbm settings go here.]] Compared to SDMs, this lets us predict richness directly 
instead of using hundreds of noisy species-level estimates, but it also discards 
any potentially-useful information regarding species turnover.

**SDMs.**
* Probably the most common approach?
* These models say that species' occurrence probabilities depend only on the values of
  the environmental variables at the time and place of the transect, not on the history
  of the environment or the history of the birds at that site
* Predict each species' occurrence probabilities individually, then add up the predictions
* [[Something about not using thresholds that throw away information]]
* Taking this approach in the face of uncertainty about observer effects would either 
  require a customized model (cite "neighborly advice" paper), or fitting [[N]] models
  times the number of species.
* To make it feasible to fit this large number of models, we used a custom wrapper 
  around the randomForest package. Random forests are flexible models that are built 
  by combining many independent sub-models ("trees") fit to different versions of the training
  data, which made it straightforward to adapt them to our situation: we simply fit one 
  tree using each of the Monte Carlo estimates of our observer effects and averaged the 
  predictions of the resulting forest.
* For the uncertainty, we used Gaussian approximations, based on the means and variances
  of the sums of independent Bernoulli random variables (cite Calabrese et al.).

**JSDM** 
* Joint species distribution model (JSDM) is a new approach that makes predictions
  about the full community composition, rather than modeling it species-by-species.
* In stacked single-species SDMs, the species-level occurrences are treated as 
  independent coin-flips whose proability depends only on the environment.
* JSDMs remove this independence assumption, and explicitly account for the
  possibility that a site will be much more (or less) suitable for birds in 
  general than one would expect based on a few environmental measurements alone.
* As a result, JSDMs do a better job of representing our uncertainty about 
  richness, while stacked SDMs underestimate it (Harris 2015).
* We used the `mistnet` package (Harris 2015) because it is the only JSDM that
  describes species' environmental associations with nonlinear functions.
* The mistnet package doesn't "know" about time series or spatial autocorrelation,
  so it did not share information among the repeated runs of the same transect.

### Ensembles
* Ensembles. Averaging predictions from multiple models tends to reduce the 
  noise in the estimates, and can thus lead to better predictions. Choosing the
  weights for the ensemble isn't straightforward because our estimates of model 
  error on the training set are biased in favor of models that
  overfit. [[Another option is to spatially cross-validate on the training set]]

### Model evaluation

* Evaluated each year between  `r settings$last_train_year` and `r settings$end_yr`
* When a model was run multiple times (e.g. once per MCMC sample for the observer
  estimates), we used the mean of those runs' point estimates as our final point 
  estimate and we calculated the uncertainty using the law of total variance (i.e.
  the average of the model runs' variance, plus the variance in the point estimates).
* Metrics:
  * (R)MSE 
    * How far, on average, are the models' predictions from the true value?
    * [[RMSE has better units than MSE (species versus squared species), but MSE is
      nice because it's additive (i.e. MSE of 2 is twice as big as MSE of 1).]]
  * Calibration/bias
    * In general, or in specific situations, is the model consistently predicting
      that richness will be too high or too low?
    * Can also report this value in terms of RMSE or MSE
  * Coverage of 95% confidence intervals
  * Gaussian deviance (simultaneously rewards good point estimates, precision, 
    and coverage)

Looked at how error changed as the time horizon of forecasting lengthened
* Error will generally tend to increase
* Error might flatten out after some number of years
* Error might increase at different rates for different methods
  * One model might be better at short time scales \& worse at long time scales

## Results and conclusions

* The mixed model (which formed the basis of our "average" predictions) found 
  that 70% of the variance in richness in the training set could be explained 
  by differences among sites, and 21% could be explained by differences among 
  observers. The remaining 9% represents residual variation, where a given 
  observer might report a different number of species on different runs of the 
  same transect.
* Over the time period we analyzed, site-level richness was mostly stationary 
  once observer effecs were accounted for, which explains the success of the 
  "average" baseline.
* If site-level richness was approximately stationary, then why does error 
  increase over time? One would expect constant error under stationarity.  
  This apparent contradiction can be resolved by examining [[Figure]]: most of 
  the increase in error was due to an influx of observers whose tendencies were
  poorly identified. Looking across the rows of panel A, we see very little 
  change in MSE over time for a given type of observer, but panel B shows a 
  steady increase in the number of observers whose estimates could only be 
  identified to within +/- 5 species (or worse) because they did not make
  enough observations in different sites in the training set to determine
  their [[biases]] any more precisely.

* None of the models reliably outperformed both baselines. Depending on the 
  criterion and whether observer differences were accounted for, the best 
  model was usually "naive" or "average." The "auto.arima" and "richness-gbm" 
  models occasionally produced less error than either baseline, but their MSE 
  values still averaged substantially higher than the baselines. Of the two 
  baselines, predictions based on the long-term mean ("average") were more 
  reliable if observer effects could be accounted for, while predictions based 
  on the most-recently observed value ("naive") were better prepared to deal 
  with observer turnover. In general, the models produced confidence intervals 
  that were too narrow; [[discuss naive and possibly mistnet here]].
* The models that were fit to individual sites (i.e. the baselines and the 
  ARIMA methods) got most or all of the inter-site variation by default, 
  which made them difficult to beat. The SDMs and GBM-richness could only 
  outperform the baselines on intra-site variation, which represented a very 
  small portion of the variance; in general, the inter-site errors were large 
  enough to overpower any such gains.
* Consistent with Distler (2015, Journal of Biogeography), we found that the 
  stacked SDMs (based on random forests) and the nonlinear regression of 
  richness on environment ("gbm") produced point estimates of similar quality, 
  though SDMs were somewhat less accurate in our case. However, we found large 
  differences between the two models in their ability to accurately report 
  uncertainty; the stacked SDMs' confidence intervals were only half of their 
  nominal widths, on average, producing the worst coverage and highest deviance 
  of any method we tested.
* [[Mistnet goes here; point estimates were worse than RF but coverage was very good]]
* Temporal richness variability doesn't track the predictors we use; sites have "inertia"
  * Cite 3 papers Ethan found on consistancy of alpha even when composition changes dramatically
  * Good for absolute accuracy, little room for improvement over baseline
  * Maybe richness is too coarse
    * abundance
    * species-level
* More important to use a reasonable error model than to get the best predictors \& basis expansions
* Ensembles help/don't help (depending on results)
* Point estimate error versus coverage
* In many ways, this is the best-case scenario for SDMs:
  * large data set, known \& consistent sampling technique over a period of decades
  * "True" absences
  * extended predictor set
  * birds arguably migrate quickly enough to make equlibrium \& especially space-for-time assumptions less ridiculous
  * errors for individual species can cancel out; only asking for aggregate values
  * Counterpoint: 
    * if most variation is among sites, maybe we want site-level means for predictors rather 
      than annual values (or, why not use both?)
* Value of stacking versus predicting the quantitity of interest directly
* Need to start doing this, even if we're bad at it.
  * We've posted predictions for the next N years of BBS data, using several methods, 
    and invite others to do so as well.
* Dig around in the residuals for insights
* SDM's and time series models may not be accounting for the ecology of north 
  american birds. Most of them migrate, so whether they are observed or not at 
  the same time every year is a function of their migratory phenology. The 
  migratory patterns of several species have been correlated with NDVI (Renfrew 
  et al 2013).

## misc

SDM assumptions:
  * we've measured the predictive power of all avialable variables, other potentially important thing that aren't available are landcover, observer skill level, (probably some others)
  * Something like stationarity:
    * e.g. "systems are at equilibrium now and will be at equilibrium in future"
    * or "species track predictor variables quickly" (no lags)
    * etc
  * independent/identically distributed observations across species, years
  * Space substitutes well for time

None of the models are true forcasts as they use observed predictor variables (except the AR1).

* Need predictions at multiple time scales
* Prediction quality at short time scales might not be a good indicator of 
  long-term effectiveness

## Box 1: Ten simple rules for making and evaluating ecological forecasts

### 1. Compare multiple modeling approaches

Typically ecological forecasts use one modeling approach or a small number of
related approaches. By fitting and evaluating multiple modeling approaches we
can learn more rapidly about the best approaches for making predictions for a
given ecological quantity. This includes comparing process based and data-driven
models and comparing the accuracy of forecasts to simple baselines to determine
if the modeled forecasts are more accurate than the naive assumption that the
world is static.

### 2. Use time-series data when possible

Forecasts describe how systems are expected to change through time. While some
areas of ecological forecasting focus primarily on time-series data, others
primarily focus on using spatial models and space-for time substitutions. Using
ecological and environmental time-series data allows the consideration of actual
dynamics from both a process [@treddenick2016a] and error structure perspective.

### 3. Pay attention to uncertainty

Understanding confidence in a forecast is just as important as understanding the
average or expected outcome. Failing to account for uncertainty can result in
overconfidence in highly uncertain outcomes leading to poor decision making and
erosion of confidence in ecological forecasts. Models should explicitly include
sources of uncertainty and how the propagate through the forecast where
possible. Evaluations of forecasts should assess the accuracy of uncertainties
as well as point estimates [@dietz2017].

### 4. Use predictors related to the question

Many ecological forecasts use data that is readily available and easy to work
with. While ease of use is a reasonable consideration it is also important to
include predictor variables that are expected to relate to the ecological
quantity being forecast and dynamic time-series of predictors instead of
long-term averages. Investing time in identifying and acquiring better predictor
variables may have at least as many benefits as using more sophisticated
modeling techniques.

### 5. Assess how forecast accuracy changes with time-lag

In general the accuracy of forecasts decreases with the length of time into the
future being forecast [@petch2015]. This decay in accuracy and the potential for
different rates of decay to result in different relative model performance at
different lead times should be considered when evaluating forecasts and
comparing models.

### 6. Include an observation model

Ecological observations are influenced by both the underlying processes and how
the system is sampled. When possible forecasts should model the factors
influencing the observation of the data.

### 7. Validate using hindcasting

To evaluate the expected accuracy and uncertainty of forecasts assess the
performance of these forecasts within existing time-series data.

### 8. Publicly archive forecasts

To allow the future evaluation of the accuracy and uncertainty of forecasts the
forecast values and/or models should be archived so that they can be assessed
after new data is generated [@mcgill2012]. Enough information should be provided
to allow an unambiguous assessment of the forecast performance.

### 9. Make short-term and long-term predictions

In cases where long-term predictions are the primary goal, short-term should
also be made to accommodate the time-scales of planning and management decisions
and to allow the accuracy of the forecasts to be quickly evaluated
[@treddenick2016a].

### 10. [[?]]

Sites' biotic and abiotic environments differ in more ways than the 
few ones you happened to measure. As a result, some sites will 
consistently have more species than a regression model (or sum of 
regression models) would predict, while others willl consistently have 
fewer. Depending on the spatial scale of these unmeasured variables, 
independent site-level random effects could be sufficient for dealing 
with this, or spatially-autocorrelated random effects might be needed.

### 2017-09-04 manuscript/draft.Rmd
title/abstract

@@
 ---
+title: "Forecasting biodiversity in breeding birds using best practices"
+author:
+- David J. Harris
+- Shawn D. Taylor
+- Ethan P. White
+bibliography: refs.bib
+csl: csl.csl
+geometry: left=3.5cm, right=3.5cm, top=2.25cm, bottom=2.25cm, headheight=12pt, letterpaper
+fontsize: 12pt
+header-includes:
+    - \usepackage{times}
+    - \usepackage{setspace}
+    - \usepackage{booktabs}
+    - \doublespacing
+    - \usepackage{lineno}
+    - \linenumbers
 output:
@@
 ---
-# Moving towards large-scale data-intensive forecasting of species richness in breeding birds
-# Using best practices for forecasting biodiversity changes in breeding birds
-Authors: Harris, Taylor and White?
 
-`r settings = yaml::yaml.load_file("settings.yaml")`
+```{r, echo=FALSE, message=FALSE}
+library(tidyverse, quietly = TRUE)
+library(kableExtra, quietly = TRUE)
+```
+
+`r settings = yaml::yaml.load_file("../settings.yaml")` 
+`r timeframe = settings$timeframes$train_22` 
+`r numbers = yaml::yaml.load_file("numbers.yaml")`
+
+## Abstract
+
+Biodiversity forecasts are important for conservation, management, and
+evaluating how well current models characterize natural systems. While the
+number of forecasts for biodiversity is increasing, there is little evidence
+showing that these forecasts work well at decadal time scales. Most biodiversity
+forecasts are not evaluated to determine how well they predict future diversity,
+fail to account for uncertainty, and do not use time-series data that captures
+the actual dynamics being studied. In this study, we used best practices and
+32-year time series to evaluate six different modeling approaches to predicting
+changes in North American breeding birds' species richness. Hindcasts for each
+method were evaluated annually for a decade at `r numbers$N_predict_sites` sites
+distributed throughout the continental United States. While each model could
+explain most of the variance in richness, none of them outperformed baseline
+models that predicted constant richness at each site. In particular, we found no
+evidence that current methods (such as species distribution models) can
+successfully turn spatial data into useful temporal predictions about
+biodiversity at decadal time-scales. To facilitate the rapid improvement of
+biodiversity forecasts, we emphasize the value of specific best practices at
+multiple stages of the forecasting pipeline, from the choice of predictor
+variables through the choice of approach for evaluating model performance over
+time.
 
@@
 Forecasting the future state of ecological systems is increasingly important for
-planning, management, and evaluating how well ecological models capture the key
-ecological processes governing a system [@clark2001, @dietz2017]. One area of
-particular interest for forecasting is biodiversity since it is important for
-ecosystem function, central to conservation planning, and expected to be
-influenced by anthropogenic global change. Anticipating potential changes in
-biodiversity is crucial for large scale management and conservation, and for
-addressing debates regarding whether local scale diversity is declining and may
-continue to decline in response to anthropogenic influences [@dornelas2014].
+planning and management, and also for quantitatively evaluating how well
+ecological models capture the key processes governing natural systems
+[@clark2001; @houlahan2017; @dietze2017]. Forecasts regarding biodiversity are
+especially important, due to biodiversity's central role in conservation
+planning and its sensitivity to anthropogenic effects [@cardinale2012;
+@diaz2015; @tilman2017]. High-profile studies forecasting large biodiversity
+declines over the coming decades have played a large role in shaping ecologists'
+priorities [as well as those of policymakers; e.g. @ipcc2014], but it is
+inherently difficult to evaluate such long-term predictions before the projected
+biodiversity declines have occurred.
 
@@
 more generally, have focused primarily on building species distributions models
-to quantify the spatial relationships between the occurrence of individual
-species and the environment. Forecasts for future environmental conditions are
-then used to predict where each species will occur in the future and the
-probabilities of occurrence are summed across species to predict future species
-richness [e.g., @thomas2004]. Alternatively, models that directly relate spatial
-patterns of species richness to the environment have been developed and
-generally perform equivalently to species distribution modeling based methods
-[@algar2009, @distler2015].
+[SDMs; @thomas2004; @thuiller2011; @urban2015]. In general, these models
+describe individual species' occurrence patterns as functions of the
+environment. Given forecasts for environmental conditions, these models can 
+predict where each species will occur in the future. These species-level 
+predictions are then combined ("stacked") to generate forecasts for species 
+richness [e.g. @calabrese2014]. Alternatively, models that directly relate spatial 
+patterns of species richness to environment conditions have been developed and 
+generally perform equivalently to stacked SDMs [@algar2009; @distler2015]. 
+This approach is sometimes referred to as "macroecological" modeling, because it 
+models the larger-scale pattern (richness) directly [@distler2015].
 
-However, despite the emerging interest in forecasting species richness and other
-aspects of biodiversity, little is known about how effectively we can anticipate
-their dynamics. This is due in part to the long time scales over which many
-ecological forecasts are applied and the resulting difficulty in assessing
-whether the predicted changes occurred. What we do know comes from a small
-number of hindcasting studies, where models are built using data on species
-occurrence and richness from the past and evaluated on their ability to predict
-contemporary patterns [e.g., @algar2009, @distler2015]. These studies are a
-valuable first step, but lack several components that are important for
-developing good forecasting models and understanding how accurately these models
-can predict the future. These components of good forecasting and evaluation (Box
-1) broadly involve: 1) expanding the use of data to include biological and
-environmental time-series [@treddenick2016]; 2) accounting for uncertainty in
-observations, processes, model choice, and forecast evaluation
-[@clark2001, @dietz2017]; and 3) conducting meaningful evaluations of the
-forecasts by hindcasting, archiving short-term forecasts, and comparing
-forecasts to baselines to determine if the forecasts are more accurate than
-assuming the system is basically static [@Perretti2013].
+Despite the emerging interest in forecasting species richness and other aspects
+of biodiversity [@jetz2007; @thuiller2011], little is known about how
+effectively we can anticipate these dynamics. This is due in part to the long
+time scales over which many ecological forecasts are applied [and the resulting
+difficulty in assessing whether the predicted changes occurred; @dietze2016].
+What we do know comes from a small number of hindcasting studies, where models
+are built using data on species occurrence and richness from the past and
+evaluated on their ability to predict contemporary patterns [e.g., @algar2009;
+@distler2015]. These studies are a valuable first step, but lack several
+components that are important for developing forecasting models with high
+predictive accuracy, and for understanding how well different methods can
+predict the future. These "best practices" for effective forecasting and
+evaluation (Box 1) broadly involve: 1) expanding the use of data to include
+biological and environmental time-series [@tredennick2016]; 2) accounting for
+uncertainty in observations and processes, [@harris2015; @yu2010]; and 3)
+conducting meaningful evaluations of the forecasts by hindcasting, archiving
+short-term forecasts, and comparing forecasts to baselines to determine whether
+the forecasts are more accurate than assuming the system is basically static
+[@perretti2013].
 
-In this paper we attempt to forecast the species richness of breeding birds at
-hundreds of locations throughout North America while following the best
-practices in Box 1 for ecological forecasting. To do this we combine 30 years of
-time-series data on bird distributions with monthly time-series of climate data
-and satellite-based remote-sensing. This 30 year time-scale has only recently
-become possible for large-scale time-series based forecasting and allows us to
+In this paper, we attempt to forecast the species richness of breeding birds at
+over 1,200 of sites located throughout North America, while following best
+practices for ecological forecasting (Box 1). To do this, we combine 32 years of
+time-series data on bird distributions from annual surveys with monthly
+time-series of climate data and satellite-based remote-sensing. Datasets that
+span a time scale of 30 years or more have only recently become available for
+large-scale time-series based forecasting. A dataset of this size allows us to
 model and assess changes a decade or more into the future in the presence of
@@
 compare traditional distribution modeling based approaches to spatial models of
-species richness, time-series methods, and simple baselines. All forecasts model
-uncertainty and observation, are evaluated across different time lags using
-hindcasting, and are publicly archived to allow future assessment. We discuss
-the implications of these practices for our understanding of, and confidence in,
-the resulting forecasts, and how we can continue to improve on these approaches
-in the future.
+species richness, time-series methods, and two simple baselines that predict
+constant richness for each site, on average (Figure 1). All of our forecasting
+models account for uncertainty and observation error, are evaluated across
+different time lags using hindcasting, and are publicly archived to allow future
+assessment. We discuss the implications of these practices for our understanding
+of, and confidence in, the resulting forecasts, and how we can continue to build
+on these approaches to improve ecological forecasting in the future.
+
+![Example predictions from six forecasting models for a single site. Data from
+1982 through 2003, connected by solid lines, were used for training the models;
+the remaining points were used for evaluating the models' forecasts. In each
+panel, point estimates for each year are shown with lines; the darker ribbon
+indicates the 68% prediction interval (1 standard deviation of uncertainty), and
+the lighter ribbon indicates the 95% prediction interval. **A.** Single-site
+models were trained independently on each site's observed richness values. The
+first two models ("average" and "naive") served as baselines. **B.** The
+environmental models were trained to predict richness based on elevation,
+climate, and NDVI; the environmental models' predictions change from year to
+year as environmental conditions change.](../figures/model_predictions.png)
 
@@
 
+We evaluated 6 types of forecasting models (Table 1) by dividing the 32 years
+of data into 22 years of training data and 10 years of data for 
+evaluating forecasts using hindcasting. We also made long term forecasts by using
+the full data set for training and making forecasts through the year 2050. For
+both time scales, we made forecasts using each model with and without
+correcting for observer effects, as described below.
+
 ### Data
 
-* BBS + retriever
-* Predictor variables
-  * Monthly climate data from interpolated weather stations (PRISM), converted
-    to BIOCLIM variables. Many of these variables were tightly correlated, so
-    we only included the 8 variables selected by Harris's (2015) BBS analysis
-  * Weather/climate and NDVI were each calculated separately for each year,
-    rather than averaged over the full multi-decade period (as Worldclim does)
-  * NDVI (GIMMS)
-  * Altitude (SRTM 90m Digital Elevation Database via raster::getData)
-* Data summary: 
-  * Number of sites, species, unique observers
-  * central moments of richness (mean about 51, sd about 12 spp)
-  * The within-site variation was much smaller (sd about 5 spp), making the
-    site-level distributions underdispersed.  
-    * This underdispersion implies
-      that common count models such as binomial, Poisson, negative binomial 
-      won't fit well.
-    * For this reason (and because the time series methods assumed 
-      Gaussian errors), we defined model performance throughout this paper in 
-      terms of Gaussians (e.g. mean squared error), rather than with count
-      distributions.
-        * Some of the modeling approaches (SDMs and JSDMs) did use discrete
-          distributions internally, however.
-* Splitting the data temporally
-  * Our data set began in `r settings$start_yr`, which was the first year that 
-    NDVI data was available from GIMMS
-  * All the models were trained on the data that was collected through 
-    `r settings$last_train_year`
-  * Sites were included in the analysis if there were at least 
-    `r settings$min_num_yrs` years of data available during this period.
-  * The models were tested/evaluated on the data from 
-    `r settings$last_train_year + 1` `r settings$end_yr`.
+**Richness data.** Bird species richness was obtained from the North American
+Breeding Bird Survey (BBS) [@pardieck2017] using the Data Retriever Python
+package [@morris2013] and rdataretriever R package [@rdataretriever]. The BBS
+data was filtered to exclude all nocturnal, cepuscular, and aquatic species
+[since these species are not well sampled by BBS methods; @hurlbert2005], as
+well as unidentified species, and hybrids. All data from surveys that did not
+meet BBS quality criteria were also excluded.
 
-### Accounting for site \& observer effects
-* Observer effects are inherent in large data sets taken by different 
-  observers, and are known to occur in BBS [citation].
-* While these biases are known to exist, they are difficult to quantify: even if
-  the different observers ran the same transects under identical conditions (which 
-  they don't), we would still only have an estimate of their differences, not an 
-  exact value.
-* We estimated the observer effects, along with our uncertainty about those
-  effects' size & direction, with a linear mixed model.
-* This model partitions the variance in observed richness values into site-level
-  variance, observer-level variance, and residual variance (e.g. variation within
-  a site from year to year or from morning to morning).
-  * No fixed effects in this model (e.g. environmental predictor variables because 
-    we're not interested in explaining differences among sites at this point; we 
-    just want to account for their existence while estimating observer effects.
-  * Subtracting off the estimated observer effect provides an estimate of how many
-    would have been found by a "typical" observer on that day; in this sense, it 
-    enables us to correct for observer differences. We then used this "corrected" 
-    version of the data for our time-series models such as the AR1, etc.
-* A simple model like this one can't tell us exactly how each observer differed from
-  the others, so it is important to represent our uncertainty about these differences.
-  Here, we did so by collecting [[N]] Monte Carlo samples from the model's posterior
-  distribution, using Stan. This provided us with [[N]] plausible values for each 
-  observer, so we did not have to commit to a single value.
-* While some methods for predicting richness might be able to use the uncertain
-  estimates of the observer effects directly, we take the simpler approach of just
-  fitting each of the downstream models with each of the Monte Carlo samples. This
-  increases the amount of computational work by a factor of [[N]], but CPU time
-  wasn't our limiting resource.
+We used observed richness values from 1982 (the first year of complete
+environmental data) to 2003 to train the models, and from 2004 to 2013 to test
+their performance. We only used BBS routes from the continental United States
+(i.e. routes where climate data was available @prism), and we restricted the
+analysis to routes that were sampled during 70% of the years in the training
+period (i.e., routes with at least 16 annual observations). The resulting
+dataset included `r numbers$N_runs` annual surveys of `r numbers$N_sites` unique
+sites, and included `r numbers$N_species` species.  Site-level richness varied
+from `r numbers$richness_summary$min` to `r numbers$richness_summary$max` with
+an average richness of `r numbers$richness_summary$mean` species.
 
-### Site-level models
+**Past environmental data.** Environmental data included a combination of
+elevation, bioclimatic variables and a remotely sensed vegetation index (the
+normalized difference vegetation index; NDVI), all of which are known to
+influence richness and distribution in the BBS data [@kent2014]. For each year
+in the dataset, we used the 4 km resolution PRISM data [@prism] to calculate
+eight bioclimatic variables identified as relevant to bird distributions
+[@harris2015]: mean diurnal range, isothermality, max temperature of the warmest
+month, mean temperature of the wettest quarter, mean temperature of the driest
+quarter, precipitation seasonality, precipitation of the wettest quarter, and
+precipitation of the warmest quarter. Satellite-derived NDVI, a primary
+correlate of richness in BBS data [@hurlbert2002], was obtained from the NDIV3g
+dataset with an 8 km resolution [@pinzon2014] and was available from 1981-2013.
+Average summer (May, June, July) and winter (December, January, Feburary) NDVI
+values were used as predictors. Elevation was from the SRTM 90m elevation
+dataset [@jarvis2008] obtained using the R package raster [@raster].
+Because BBS routes are 40-km transects rather than point counts, we used the
+average value of each environmental variable within a 40 km radius of each BBS
+route's starting point.
 
-[[note that these all use a Gaussian error distribution]]
+**Future environmental projections.** We made long term forecasts from 2014-2050
+using the CMIP5 multi-model ensemble dataset as the source for climate
+variables [@brekke2013]. Precipitation and temperature from 37 downscaled model
+runs [@brekke2013, see Table S1] using the RCP6.0 scenario were averaged
+together to create a single ensemble used to calculate the bioclimatic variables
+for North America. For NDVI we used the per-site average values from 2000-2013
+as a simple forecast. For observer effects (see below) each site was set to have
+zero observer bias.
 
-**Baseline models.** The two simplest models we fit treated site-level richness
-observations either as uncorrelated noise around a site-level constant (the 
-"average" model) or as an autoregressive model with a single year of history 
-(the "naive" model). The predictions from these two models are qualitatively 
-different: predictions from the "average" model are centered on the average 
-richness and the confidence intervals are narrow and constant-width. The
-"naive" model, in contrast, predicts that future observations will be similar to
-the final observed value (e.g. the value observed in `$r settings$last_train_year`), 
-and the confidence intervals expand rapidly as the predictions extend farther into 
-the future.  Each of these model types was fit separately to each BBS route in two
-ways: first, we fit the model to the raw richness values, without accounting for
-possible differences among observers at the site.  Then, we fit the models to the
-residuals from our observer model (i.e. fitting the model to an estimate of how
-many species a typical observer would have found, rather than to the raw data). 
-In the latter case, we averaged across [[N]] models that were fit separately to 
-the [[N]] Monte Carlo estimates of the observer effects, to account for our
-uncertainty in the true values of those effects.
+### Accounting for observer effects
 
-**Time series models.** Like the two baseline models, these models only looked at 
-one site at a time, but they could include additional complexity. As above, these
-models were trained both with the raw richness time series and with Monte Carlo
-estimates of the observer-corrected time series. The  Auto-ARIMA models (based on 
-Hyndman's 2016 `auto.arima` function) could include an autoregressive component 
-(as in the "naive" model, but with the possibility of longer-term dependencies
-in the underlying process), a moving average component (where the noise can have 
-serial autocorrelation) and an integration/differencing component (so that the 
-analysis could be performed on sequential differences of the raw data, rather than
-on the observed time series itself). The `auto.arima` function chooses whether to 
-include each of these components, and how many terms to include for each one, using
-AICc, as described in Hyndman (2016). We used the default settings for this function,
-apart from setting `seasonal = FALSE` because our time series does not include a 
-seasonal component.  [[Note: should probably record somewhere what order model
-the auto.arima function tends to choose.  Eyeballing the results, seems to be most 
-similar to the naive model, but would be good to check.]] We also fit an 
-"Auto-ARIMA + environment" model, in which each site-level time series could 
-also be predicted by annual changes in our environmental variables at that site.
+Observer effects are inherent in large data sets collected by different
+observers, and are known to occur in BBS [@sauer1994]. For each forecasting
+approach, we trained two versions of the corresponding model: one with
+corrections for differences among observers, and one without (Figure 2). We
+estimated the observer effects (and associated uncertainty about those effects)
+with a linear mixed model, with observer as a random effect, built in the Stan
+probabilistic programming language [@carpenter2017]. Because observer and site are
+strongly related (observers tend to repeatedly sample the same site), site was
+also included as a random effect to ensure that inferred deviations were
+actually observer-related (as opposed to being related to the sites that a given
+observer happened to see). The resulting model partitions the variance in
+observed richness values into site-level variance, observer-level variance, and
+residual variance (e.g.  variation within a site from year to year). The
+site-level estimates can also be used directly as the "average" baseline model
+(see below). The estimated observer effects can be subtracted from the richness
+values for a particular observer to provide an estimate of how many species
+would have been found by a "typical" observer. To incorporate uncertainty in
+these "corrected" richness values into the forecasting models we collected 500
+Monte Carlo samples from the model's posterior distribution, and fit each of the
+downstream models with each of the Monte Carlo samples. Each Monte Carlo sample
+represented a different possible set of observer-level and site-level random
+effect values across the full 32-year dataset.
 
-### Continental-scale models
+![**A.** Model predictions for Pennsylvania route 35 when all observers are
+treated the same (black points). **B.** Model predictions for the same route
+when accounting for systematic differences between observers (represented by the
+points' colors). In this example most models are made more robust to observer
+turnover. Note that the "naive" model is less sensitive to observer turnover,
+and does not benefit as much from modeling
+it.](../figures/observer_predictions.png)
 
-So far, each model has been fit to one site at a time, with no information 
-(apart from observer effects) shared among sites. When continental-scale data 
-is available, ecologists often use correlative models to predict richness 
-changes over time from environmental data. Here, we tested three variants of 
-this approach. Here, we did *not* include the site-level random effects as 
-predictors, meaning that this approach implicitly assumes that two sites 
-with identical Bioclim, elevation, and NDVI values should have identical 
-richness distributions. As above, we included observer effects & the 
-associated uncertainty by running this model [[N]] times (once per MCMC 
-sample).
+### Models: site-level models
 
-**"Macroecological" model.** The most straightforward way to predict richness is 
-to fit a richness model. We chose the `gbm` package for fitting boosted 
-regression trees as our example of such a model. Boosted regression trees is a 
-common approach in the species distribution modeling literature [refs] and 
-works by fitting thousands of small tree-structured models sequentially, with
-each tree optimized to reduce the error of its predecessors. For the reasons 
-discussed above, this model was optimized using a Gaussian likelihood. [[other 
-gbm settings go here.]] Compared to SDMs, this lets us predict richness directly 
-instead of using hundreds of noisy species-level estimates, but it also discards 
-any potentially-useful information regarding species turnover.
+Three of the models used in this study were fit to each site separately, with no
+environmental information (Table 1). These models were fit to each BBS route
+twice: once using the residuals from the observer model, and once using the raw
+richness values. When correcting for observer effects, we averaged across 500
+models that were fit separately to the 500 Monte Carlo estimates of the observer
+effects, to account for our uncertainty in the true values of those effects. All
+of these models use a Gaussian error distribution (rather than a count
+distribution) for reasons discussed below (see "Model evaluation").
 
-**SDMs.**
-* Probably the most common approach?
-* These models say that species' occurrence probabilities depend only on the values of
-  the environmental variables at the time and place of the transect, not on the history
-  of the environment or the history of the birds at that site
-* Predict each species' occurrence probabilities individually, then add up the predictions
-* [[Something about not using thresholds that throw away information]]
-* Taking this approach in the face of uncertainty about observer effects would either 
-  require a customized model (cite "neighborly advice" paper), or fitting [[N]] models
-  times the number of species.
-* To make it feasible to fit this large number of models, we used a custom wrapper 
-  around the randomForest package. Random forests are flexible models that are built 
-  by combining many independent sub-models ("trees") fit to different versions of the training
-  data, which made it straightforward to adapt them to our situation: we simply fit one 
-  tree using each of the Monte Carlo estimates of our observer effects and averaged the 
-  predictions of the resulting forest.
-* For the uncertainty, we used Gaussian approximations, based on the means and variances
-  of the sums of independent Bernoulli random variables (cite Calabrese et al.).
+\begin{singlespace}
 
-**JSDM** 
-* Joint species distribution model (JSDM) is a new approach that makes predictions
-  about the full community composition, rather than modeling it species-by-species.
-* In stacked single-species SDMs, the species-level occurrences are treated as 
-  independent coin-flips whose proability depends only on the environment.
-* JSDMs remove this independence assumption, and explicitly account for the
-  possibility that a site will be much more (or less) suitable for birds in 
-  general than one would expect based on a few environmental measurements alone.
-* As a result, JSDMs do a better job of representing our uncertainty about 
-  richness, while stacked SDMs underestimate it (Harris 2015).
-* We used the `mistnet` package (Harris 2015) because it is the only JSDM that
-  describes species' environmental associations with nonlinear functions.
-* The mistnet package doesn't "know" about time series or spatial autocorrelation,
-  so it did not share information among the repeated runs of the same transect.
+```{r table, echo=FALSE}
 
-### Ensembles
-* Ensembles. Averaging predictions from multiple models tends to reduce the 
-  noise in the estimates, and can thus lead to better predictions. Choosing the
-  weights for the ensemble isn't straightforward because our estimates of model 
-  error on the training set are biased in favor of models that
-  overfit. [[Another option is to spatially cross-validate on the training set]]
+table_caption = "Six forecasting models. Single-site models were trained
+site-by-site, without environmental data. Environmental models were trained
+using all sites together, without information regarding which transects occurred
+at which site or during which year. Most of the models were trained to predict
+richness directly. This mirrors the standard application of these
+techniques. Separate random forest SDMs were fit for each species and used to
+predict the probability of that species occurring at each site. The
+species-level probabilities at a site were summed to predict richness. The
+mistnet JSDM was trained to predict the full species composition at each site,
+and the number of species in its predictions was used as an estimate of
+richness."
+
+table_data = read_csv("../figures/table.csv",col_types = "ccccc")
+
+table_data %>%
+  knitr::kable(
+    format = "latex",
+    escape = FALSE,
+    booktabs = TRUE,
+    caption = table_caption
+  ) %>%
+  add_header_above(c(" " = 2, "Predictors" = 3)) %>%
+  group_rows("Single-site models", 1, 3) %>%
+  group_rows("Environmental models", 4, 6) %>% kable_styling(position = "center")
+```
+
+\end{singlespace}
+
+**Baseline models.** We used two simple baseline models as a basis for
+comparison with the more complex models (Figure 2A). These baselines treated
+site-level richness observations either as uncorrelated noise around a
+site-level constant (the "average" model) or as an autoregressive model with a
+single year of history [the "naive" model, @hyndman2014]. Predictions from the
+"average" model are centered on the average richness observed during training,
+and the confidence intervals are narrow and constant-width. The "naive" model,
+in contrast, predicts that future observations will be similar to the final
+observed value (e.g., in our hindcasts the value observed in `r
+timeframe$last_train_year`), and the confidence intervals expand rapidly as the
+predictions extend farther into the future. Both models' richness predictions
+are centered on a constant value, so neither model can anticipate any trends in
+richness or any responses to future environmental changes.
+
+**Time series models.** We used Auto-ARIMA models [based on the `auto.arima`
+function in @forecast1] to represent an array of different time-series modeling
+approaches. These models can include an autoregressive component (as in the
+"naive" model, but with the possibility of longer-term dependencies in the
+underlying process), a moving average component (where the noise can have serial
+autocorrelation) and an integration/differencing component (so that the analysis
+could be performed on sequential differences of the raw data, accommodating more
+complex patterns including trends). The `auto.arima` function chooses whether to
+include each of these components (and how many terms to include for each one)
+using AICc [@forecast1]. Since there is no seasonal component to the BBS
+time-series, we did not include a season component in these models. Otherwise we
+used the default settings for this function [@forecast1].
+
+### Models: environmental models
+
+In contrast to the single-site models, most attempts to predict species richness
+focus on using correlative models based on environmental variables. We tested
+three common variants of this approach: direct modeling of species richness;
+stacking individual species distribution models; and joint species distribution
+models (JSDMs). Following the standard approach, site-level random effects were
+not included in these models as predictors, meaning that this approach
+implicitly assumes that two sites with identical Bioclim, elevation, and NDVI
+values should have identical richness distributions. As above, we included
+observer effects and the associated uncertainty by running these models 500
+times (once per MCMC sample).
+
+**"Macroecological" model: richness GBM.** We used a boosted regression tree
+model using the `gbm` package [Ridgeway *et al.* -@gbm] to directly model
+species richness as a function of environmental variables. Boosted regression
+trees are a form of tree-based modeling that work by fitting thousands of small
+tree-structured models sequentially, with each tree optimized to reduce the
+error of its predecessors. They are flexible models that are considered well
+suited for prediction [@elith2008]. This model was optimized using a Gaussian
+likelihood, with a maximum interaction depth of 5, shrinkage of 0.015, and up to
+10,000 trees. The number of trees used for prediction was selected using the
+"out of bag" estimator; this number averaged `r numbers$n_gbm[1]` for the
+non-observer data and `r numbers$n_gbm[2]` for the observer-corrected data.
+
+**Species Distribution Model: stacked random forests.** Species distribution
+models (SDMs) predict individual species' occurrence probabilities using
+environmental variables. Species-level models are used to predict richness by
+summing the predicted probability of occupancy across all species at a site.
+This avoids known problems with the use of thresholds for determining whether or
+not a species will be present at a site [@calabrese2014; @pellissier2013].
+Following @calabrese2014, we calculated the uncertainty in our richness estimate
+by treating richness as a sum over independent Bernoulli random variables:
+$\sigma^2_{richness} = \sum_i{p_i (1 - p_i)}$, where $i$ indexes species. By
+itself, this approach is known to underestimate the true community-level
+uncertainty because it ignores the uncertainty in the species-level probabilites
+[@calabrese2014]. To mitigate this problem, we used an ensemble of 500 estimates
+for each of the species-level probabilities instead of just one, propagating the
+uncertainty forward. We obtained these estimates using random forests, a common
+approach in the species distribution modeling literature. Random forests are
+constructed by fitting hundreds of independent regression trees to
+randomly-perturbed versions of the data [@cutler2007; @caruana2008]. When
+correcting for observer effects, each of the 500 trees in our species-level
+random forests used a different Monte Carlo estimate of the observer effects as
+a predictor variable.
+
+**Joint Species Distribution Model: mistnet.** Joint species distribution models
+(JSDMs) are a new approach that makes predictions about the full composition of
+a community instead of modeling each species independently as above
+[@warton2015]. JSDMs remove the assumed independence among species and
+explicitly account for the possibility that a site will be much more (or less)
+suitable for birds in general (or particular groups of birds) than one would
+expect based on the available environmental measurements alone. As a result,
+JSDMs do a better job of representing uncertainty about richness than stacked
+SDMs [@harris2015; @warton2015]. We used the `mistnet` package
+[@harris2015] because it is the only JSDM that describes species' environmental
+associations with nonlinear functions.
 
@@
 
-* Evaluated each year between  `r settings$last_train_year` and `r settings$end_yr`
-* When a model was run multiple times (e.g. once per MCMC sample for the observer
-  estimates), we used the mean of those runs' point estimates as our final point 
-  estimate and we calculated the uncertainty using the law of total variance (i.e.
-  the average of the model runs' variance, plus the variance in the point estimates).
-* Metrics:
-  * (R)MSE 
-    * How far, on average, are the models' predictions from the true value?
-    * [[RMSE has better units than MSE (species versus squared species), but MSE is
-      nice because it's additive (i.e. MSE of 2 is twice as big as MSE of 1).]]
-  * Calibration/bias
-    * In general, or in specific situations, is the model consistently predicting
-      that richness will be too high or too low?
-    * Can also report this value in terms of RMSE or MSE
-  * Coverage of 95% confidence intervals
-  * Gaussian deviance (simultaneously rewards good point estimates, precision, 
-    and coverage)
+We defined model performance for all models in terms of continuous Gaussian
+errors, instead of using discrete count distributions. Variance in species
+richness within sites was lower than predicted by several common count models,
+such as the Poisson or binomial (i.e. richness was underdispersed for individual
+sites), so these count models would have had difficulty fitting the data [cf.
+@calabrese2014]. The use of a continuous distribution is adequate here, since
+richness had a relatively large mean (`r numbers$richness_summary$mean`) and all
+models produce continuous richness estimates. When a model was run multiple
+times for the purpose of correcting for observer effects, we used the mean of
+those runs' point estimates as our final point estimate and we calculated the
+uncertainty using the law of total variance (i.e. the average of the model runs'
+variance, plus the variance in the point estimates).
 
-Looked at how error changed as the time horizon of forecasting lengthened
-* Error will generally tend to increase
-* Error might flatten out after some number of years
-* Error might increase at different rates for different methods
-  * One model might be better at short time scales \& worse at long time scales
+We evaluated each model's forecasts using the data for each year between `r
+timeframe$last_train_year + 1` and `r timeframe$end_yr`. We used three metrics
+for evaluating performance: 1) root-mean-square error (RMSE) to determine how
+far, on average, the models' predictions were from the observed value; 2) the
+95% prediction interval coverage to determine how well the models predicted the
+range of possible outcomes; and 3) deviance (i.e. negative 2 times the Gaussian
+log-likelihood) as an integrative measure of fit incorporating good point
+estimates, precision, and coverage. In addition to evaluating forecast
+performance in general, we evaluated how performance changed as the time horizon
+of forecasting increased by plotting performance metrics against year. Finally,
+we decomposed each model's squared error into two components: the squared error
+associated with site-level means and the squared error associated with annual
+fluctuations in richness within a site. This decomposition describes the extent
+to which each model's error depends on consistent differences among sites versus
+changes in site-level richness from year to year.
 
-## Results and conclusions
+All analyses were conducted using R [@R]. Primary R packages used in the
+analysis included dplyr [@dplyr], tidyr [@tidyr], gimms [@gimms], sp [@sp1;
+@sp2], raster [@raster], prism [@prism], rdataretriever [@rdataretriever],
+forecast [@forecast1; @forecast2], git2r [@git2r], ggplot [@ggplot2], mistnet
+[@harris2015], viridis [@viridis], rstan [@rstan], yaml [@yaml], purrr [@purrr],
+gbm [Ridgeway *et al.* -@gbm], randomForest [@randomForest]. Code to fully
+reproduce this analysis is available on GitHub
+(https://github.com/weecology/bbs-forecasting) and archived on Zenodo
+(https://doi.org/10.5281/zenodo.839581).
 
-* The mixed model (which formed the basis of our "average" predictions) found 
-  that 70% of the variance in richness in the training set could be explained 
-  by differences among sites, and 21% could be explained by differences among 
-  observers. The remaining 9% represents residual variation, where a given 
-  observer might report a different number of species on different runs of the 
-  same transect.
-* Over the time period we analyzed, site-level richness was mostly stationary 
-  once observer effecs were accounted for, which explains the success of the 
-  "average" baseline.
-* If site-level richness was approximately stationary, then why does error 
-  increase over time? One would expect constant error under stationarity.  
-  This apparent contradiction can be resolved by examining [[Figure]]: most of 
-  the increase in error was due to an influx of observers whose tendencies were
-  poorly identified. Looking across the rows of panel A, we see very little 
-  change in MSE over time for a given type of observer, but panel B shows a 
-  steady increase in the number of observers whose estimates could only be 
-  identified to within +/- 5 species (or worse) because they did not make
-  enough observations in different sites in the training set to determine
-  their [[biases]] any more precisely.
+## Results
 
-* None of the models reliably outperformed both baselines. Depending on the 
-  criterion and whether observer differences were accounted for, the best 
-  model was usually "naive" or "average." The "auto.arima" and "richness-gbm" 
-  models occasionally produced less error than either baseline, but their MSE 
-  values still averaged substantially higher than the baselines. Of the two 
-  baselines, predictions based on the long-term mean ("average") were more 
-  reliable if observer effects could be accounted for, while predictions based 
-  on the most-recently observed value ("naive") were better prepared to deal 
-  with observer turnover. In general, the models produced confidence intervals 
-  that were too narrow; [[discuss naive and possibly mistnet here]].
-* The models that were fit to individual sites (i.e. the baselines and the 
-  ARIMA methods) got most or all of the inter-site variation by default, 
-  which made them difficult to beat. The SDMs and GBM-richness could only 
-  outperform the baselines on intra-site variation, which represented a very 
-  small portion of the variance; in general, the inter-site errors were large 
-  enough to overpower any such gains.
-* Consistent with Distler (2015, Journal of Biogeography), we found that the 
-  stacked SDMs (based on random forests) and the nonlinear regression of 
-  richness on environment ("gbm") produced point estimates of similar quality, 
-  though SDMs were somewhat less accurate in our case. However, we found large 
-  differences between the two models in their ability to accurately report 
-  uncertainty; the stacked SDMs' confidence intervals were only half of their 
-  nominal widths, on average, producing the worst coverage and highest deviance 
-  of any method we tested.
-* [[Mistnet goes here; point estimates were worse than RF but coverage was very good]]
-* Temporal richness variability doesn't track the predictors we use; sites have "inertia"
-  * Cite 3 papers Ethan found on consistancy of alpha even when composition changes dramatically
-  * Good for absolute accuracy, little room for improvement over baseline
-  * Maybe richness is too coarse
-    * abundance
-    * species-level
-* More important to use a reasonable error model than to get the best predictors \& basis expansions
-* Ensembles help/don't help (depending on results)
-* Point estimate error versus coverage
-* In many ways, this is the best-case scenario for SDMs:
-  * large data set, known \& consistent sampling technique over a period of decades
-  * "True" absences
-  * extended predictor set
-  * birds arguably migrate quickly enough to make equlibrium \& especially space-for-time assumptions less ridiculous
-  * errors for individual species can cancel out; only asking for aggregate values
-  * Counterpoint: 
-    * if most variation is among sites, maybe we want site-level means for predictors rather 
-      than annual values (or, why not use both?)
-* Value of stacking versus predicting the quantitity of interest directly
-* Need to start doing this, even if we're bad at it.
-  * We've posted predictions for the next N years of BBS data, using several methods, 
-    and invite others to do so as well.
-* Dig around in the residuals for insights
-* SDM's and time series models may not be accounting for the ecology of north 
-  american birds. Most of them migrate, so whether they are observed or not at 
-  the same time every year is a function of their migratory phenology. The 
-  migratory patterns of several species have been correlated with NDVI (Renfrew 
-  et al 2013).
+The site-observer mixed model found that 70% of the variance in richness in the
+training set could be explained by differences among sites, and 21% could be
+explained by differences among observers. The remaining 9% represents residual
+variation, where a given observer might report a different number of species in
+different years. In the training set, the residuals had a standard deviation of
+about `r numbers$resid_sd` species. After correcting for observer differences,
+there was little temporal autocorrelation in these residuals (i.e. the residuals
+in one year explain `r round(100 *numbers$residual_autocor^2, 1)`% of the
+variance in the residuals of the following year), suggesting that richness was
+approximately stationary between 1982 and 2003.
 
-## misc
+When comparing forecasts for richness across sites all methods performed well
+(Figure 3; all $R^2 > 0.5$). However SDMs (both stacked and joint) and the
+macroecological model all failed to successfully forecast the highest-richness
+sites, resulting in a notable clustering of predicted values near ~60 species
+and the poorest model performance ($R^2$=`r numbers$env_R2s`, versus $R^2$=`r
+numbers$ts_R2s` for the within-site methods).
 
-SDM assumptions:
-  * we've measured the predictive power of all avialable variables, other potentially important thing that aren't available are landcover, observer skill level, (probably some others)
-  * Something like stationarity:
-    * e.g. "systems are at equilibrium now and will be at equilibrium in future"
-    * or "species track predictor variables quickly" (no lags)
-    * etc
-  * independent/identically distributed observations across species, years
-  * Space substitutes well for time
+![Performance of six forecasting models for predicting species richness one year
+(2004) and ten years into the future (2013). Plots show observed vs. predicted
+values for species richness. Models were trained with data from 1982-2003.  In
+general, the single-site models (**A**) outperformed the environmental models
+(**B**). The accuracy of the predictions generally declined as the timescale of
+the forecast was extended from 2004 to 2013.](../figures/scatter.png)
 
-None of the models are true forcasts as they use observed predictor variables (except the AR1).
+While all models generally performed well in absolute terms (Figure 3), none
+consistently outperformed the "average" baseline (Figure 4). The auto-ARIMA was
+generally the best-performing non-baseline model, but in many cases (67% of the
+time), the auto.arima procedure selected a model with only an intercept term
+(i.e. no autoregressive terms, no drift, and no moving average terms), making it
+similar to the "average" model. All five alternatives to the "average" model
+achieved lower error on some of the sites in some years, but each one had a
+higher mean absolute error and higher mean deviance (Figure 4).
 
-* Need predictions at multiple time scales
-* Prediction quality at short time scales might not be a good indicator of 
-  long-term effectiveness
+![Difference between the forecast error of models and the error of the average
+baseline using both absolute error (**A.**) and deviance (**B.**). Differences
+are taken for each site and testing year so that errors for the same forecast
+are directly compared. The error of the average baseline is by definition zero
+and is indicated by the horizontal gray line.  None of the five models provided
+a consistent improvement over the average baseline. The absolute error of the
+models was generally similar or larger than that of the "average" model, with
+large outliers in both directions.  The deviance of the models was also
+generally higher than the "average" baseline.](../figures/model_violins.png)
 
-## Box 1: Ten simple rules for making and evaluating ecological forecasts
+![Change in performance of the six forecasting models with the time scale of the
+forecast (1-10 years into the future). **A.** Root mean square error (rmse; the
+error in the point estimates) shows the three environmental models tending to
+show the largest errors at all time scales and the models getting worse as they
+forecast further into the future at approximately the same rate. **B.** Deviance
+(lack of fit of the entire predictive distribution) shows the stacked species
+distribution models with much higher error than other models and shows that the
+"naive" model's deviance grows relatively quickly. **C.** Coverage of a model's
+95% confidence intervals (how often the observed values fall inside the
+predicted range; the black line indicates ideal performance) shows that the
+"naive" model's predictive distribution is too wide (capturing almost all of the
+data) and the stacked SDM's predictive distribution is too narrow (missing
+almost a third of the observed richness values by
+2014).](../figures/performance_time.png)
+
+Most models produced confidence intervals that were too narrow, indicating
+overconfident predictions (Figure 5C). The random forest-based SDM stack was the
+most overconfident model, with only `r numbers$rf_coverage_pct`% of observations
+falling inside its 95% confidence intervals. This stacked SDM's narrow
+predictive distribution caused it to have notably higher deviance (Figure 5B)
+than the next-worst model, even though its point estimates were not unusually
+bad in terms of RMSE (5A). As discussed elsewhere [@harris2015], this
+overconfidence is a product of the assumption in stacked SDMs that errors in the
+species-level predictions are independent. The GBM-based "macroecological" model
+and the mistnet JSDM had the best calibrated uncertainty estimates (Figure
+5B)and therefore their relative performance was higher in terms of deviance than
+in terms of RMSE. The "naive" model was the only model whose confidence
+intervals were too wide (Figure 5C), which can be attributed to the rapid rate
+at which these intervals expand (Figure 1).
+
+Partitioning each model's squared error into consistent errors across sites and
+errors that fluctuated year-to-year shows that the majority of the error comes
+from errors in the site-level mean (Figure 6). The "average" model, which was
+based on the site-level mean, had the lowest error in this regard. In contrast,
+the three environmental models showed the largest biases at the site level; this
+makes sense, given that they could not explicitly distinguish among sites with
+similar climate, NDVI, and elevation.
+
+![Partitioning of the squared error for each model into site and year
+components. The site-level mean component shows consistent over or under
+estimates of richness at a site across years. The annual fluctuation compoonent
+shows errors in predicting fluctuations in a site's richness over time. Both
+components of the mean squared error were lower for the single-site models than
+for the environmental models.](../figures/barcharts.png)
+
+Accounting for differences among observers generally improved measures of model
+fit (Figure 7). Improvements primarily resulted from a small number of forecasts
+where observer turnover caused a large shift in the reported richness values
+(Appendix [[Figure Number]]). The naive baseline was less sensitive to these
+shifts, because it largely ignored the richness values reported by observers
+that had retired by the end of the training period (Figure 1). The average
+model, which gave equal weight to observations from the whole training period,
+showed a larger decline in performance when not accounting for observer effects
+-- especially in terms of coverage. The performance of the mistnet JSDM was
+notable here, because its prediction intervals retained good coverage even when
+not correcting for observer differences, which we attribute to the JSDM's
+ability to model this variation with its latent variables.
+
+![Controlling for differences among observers generally improved each model's
+predictions, on average. The magnitude of this effect was negligible for the
+Naive baseline, however.](../figures/obs_arrows.png)
+
+## Discussion
+
+Forecasting is an emerging imperative in ecology; as such, the field needs to
+develop and follow best practices for conducting and evaluating ecological
+forecasts [@clark2001]. We have used a number of these practices (Box 1) in a
+single study that builds and evaluates forecasts of biodiversity in the form of
+species richness. The results of this effort are both promising and humbling.
+When comparing forecasts across sites, many different approaches to forecasting
+produce reasonable forecasts (Figure 3). If a site is predicted to have a high
+number of species in the future, relative to other sites, it generally does.
+However, none of the methods evaluated reliably determined how site-level
+richness changes over time (Figure 6), which is generally the stated purpose of
+these forecasts. As a result, baseline models, which did not attempt to
+anticipate changes in richness over time, generally provided the best forecasts
+for future biodiversity. While this study is restricted to breeding birds in
+North America, its results are consistent with a growing literature on the
+limits of ecological forecasting, as discussed below.
+
+The most commonly used methods for forecasting future biodiversity, SDMs and
+macroecological models, both produced worse forecasts than time-series models
+and simple baselines. This weakness suggests that predictions about future
+biodiversity change should be viewed with skepticism unless the underlying
+models have been validated temporally, via hindcasting and comparison with
+simple baselines. Since site-level richness is relatively stable, spatial
+validation is not enough: a model can have high accuracy across spatial
+gradients without being able to predict changes over time. This gap between
+spatial and temporal accuracy is known to be important for species-level
+predictions [@rapacciuolo2012; @oedekoven2017]; our results indicate that it is
+substantial for higher-level patterns like richness as well. SDMs' poor temporal
+predictions are particularly sobering, as these models have been one of the main
+foundations for estimates of the predicted loss of biodiversity to climate
+change over the past decade or so [@thomas2004; @thuiller2011; @urban2015]. Our
+results also highlight the importance of comparing multiple modeling approaches
+when conducting ecological forecasts, and in particular, the value of comparing
+results to simple baselines to avoid over-interpreting the information present
+in these forecasts [Box 1]. Disciplines that have more mature forecasting
+cultures often do this by reporting "forecast skill", i.e., the improvement in
+the forecast relative to a simple baseline [@jolliffe2003]. We recommend
+following the example of @ye2015 and adopting this approach in future ecological
+forecasting research.
+
+When comparing different methods for forecasting our results demonstrate the
+importance of considering uncertainty [Box 1; @clark2001; @dietz2016]. Previous
+comparisons between stacked SDMs and macroecological models reported that the
+methods yielded equivalent results for forecasting diversity [@algar2009;
+@distler2015]. While our results support this equivalence for point estimates,
+they also show that stacked SDMs dramatically underestimate the range of
+possible outcomes; after ten years, more than a third of the observed richness
+values fell outside the stacked SDMs' 95% prediction intervals. Consistent with
+@harris2015 and @warton2015, we found that JSDMs' wider prediction intervals
+enabled them to avoid this problem. Macroecological models appear to share this
+advantage, while being considerably easier to implement.
+
+The accuracy for forecasts is expected to decline as forecasts are made further
+into the future, so it is important to evaluate forecasts at different lead
+times [Box 1; @petchey2015]. For deviance (the error metric that evaluates both
+models' point estimates and their uncertainty simultaneously), there is some
+indication that the JSDM and the nonlinear richness model converge towards the
+time-series and baseline models as the lead time of the forecast increases. This
+potential convergence is biologically plausible because at short lead times,
+environmental changes are relatively small and changes in the biota may lag
+behind changes in the environment. However, as the lead time of the forecast
+increases the amount of environmental change should increase and the systems
+will have more time to respond, potentially resulting in shifts towards improved
+relative performance of models incorporating environmental and ecological
+information [@blonder2017]. This is potentially important for interpreting our
+results because we only evaluated annual forecasts up to a decade into the
+future. In contrast, many ecological forecasts are temporally aggregated to 5-30
+year timesteps and have lead times of up to a century. These long-term forecasts
+are difficult to evaluate given the small number of century scale datasets, but
+are important for understanding changes in biodiversity at some of the lead
+times relevant for conservation and management. Two studies have assessed models
+of species richness at longer lead times [@algar2009; @distler2015], but the
+results were not compared to baseline or time-series models (in part due to data
+limitations) making them difficult to compare to our results. Given the
+potential changes with lead time of the models and processes are most relevant
+to forecasting, it is important that more long lead time evaluation studies are
+conducted and that the inferences drawn from short-term and long-term
+forecasting studies be compared. For true forecasts, short term forecasts will
+allow for relatively rapid validation of models by researchers without needing
+to wait decades for data to accumulate [@tredennick2016; @dietze2016]. However,
+since drivers of species richness differ at different temporal scales
+[@rosenzweig1995; @white2004; @white2007], short-term forecasts should be viewed
+as informing, not replacing, long-term forecasts [@tredennick2016].
+
+While it is possible that models that include species' relationships to their
+environments or direct environmental constraints on richness will provide better
+fits at longer lead times, it is also possible that they will continue to
+produce forecasts that are worse than baselines assuming that the systems are
+static.  This would be expected to occur if richness in these systems is not
+changing over the relevant multi-decadal time scales, which would make simpler
+models with no directional change more appropriate. Recent suggestions that
+local scale richness in some systems is not changing directionally at
+multi-decadal scales supports this possibility [@brown2001; @ernest2001;
+@vellend2013; @dornelas2014]. A lack of change in richness may be expected even
+in the presence of substantial changes in environmental conditions and species
+composition at a site due to replacement of species from the regional pool
+[@brown2001; @ernest2001]. On average, the Breeding Bird Survey sites used in
+this study show little change in richness [site-level SD of `r numbers$resid_sd`
+species, after controlling for differences among observers; see also
+@lasorte2005]. The absence of rapid change in this dataset is beneficial for the
+absolute accuracy of forecasts across different sites: when a past year's
+richness is already known, it is easy to estimate future richness.  The
+site-level stability of the BBS data also explains why SDMs and macroecological
+models perform relatively well at predicting future richness, despite failing to
+capture changes in richness over time. However, this stability also makes it
+difficult to improve forecasts relative to simple baselines, since those
+baselines are already close to representing what is actually occurring in the
+system. These results suggest that single-site models should be actively
+considered for forecasts of richness and other stable aspects of
+biodiversity. Our results also suggest that future efforts to understand and
+forecast biodiversity should incorporate species composition, since lower-level
+processes are expected to be more dynamic [@ernest2001; @dornelas2014] and
+contain more useful information [@harris2015].
+
+Future biodiversity forecasting efforts also need to address the uncertainty
+introduced by the error in forecasting the environmental conditions that are
+used as predictor variables. In this, and other hindcasting studies, the
+environmental conditions for the "future" are known because the data has already
+been observed. However, in real forecasts the environmental conditions
+themselves have to be predicted, and environmental forecasts will also have
+uncertainty and bias. Ultimately, ecological forecasts that use environmental
+data will therefore be more uncertain than our current hindcasting efforts. It
+is important to correctly incorporate this uncertainty in the predictor
+variables into forecasting models [@clark2001; @dietze2017]. Limitations in
+forecasting future environmental conditions, particularly at small scales, will
+present continued challenges for models incorporating environmental variables,
+and this may result in a continued advantage for simple single-site approaches.
+
+In addition to comparing and improving the process models used for forecasting
+it is important to consider the observation models.  When working with any
+ecological dataset, there are imperfections in the sampling process that have
+the potential to influence results. With large scale surveys and citizen science
+datasets, such as the Breeding Bird Survey, these issues are potentially
+magnified by the large number of different observers and by major differences in
+the habitats and species being surveyed [@sauer1994]. Accounting for differences
+in observers reduced the average error in our point estimates and also improved
+the coverage of the confidence intervals. In addition, controlling for observer
+effects resulted in changes in which models performed best, most notably
+reducing the relative performance of the naive model's point estimates. This
+suggests that, prior to accounting for observer effects, the naive model
+performed well because it was capable of accommodating rapid shifts in estimated
+richness introduced by changes in the observer. These kinds of rapid changes
+were difficult for the other single-site models to accommodate, and so the
+average performance of the ARIMA and average models improved once this source of
+observation error was addressed. This demonstrates that modeling observation
+error can be important for properly estimating and reducing uncertainty in
+forecasts and can also lead to changes in the best methods for forecasting [Box
+1]. Another key aspect of an ideal observation model is imperfect detection. In
+this study, we did not address differences in detection probability across
+species and sites [@boulinier1998] since there is no clear way to address this
+issue using NA Breeding Bird Survey data without making strong assumptions about
+the data [i.e., assuming there is no biological variation in stops along a
+route; @white2010], but this would be a valuable addition to future forecasting
+models.
+
+The science of forecasting biodiversity remains in its infancy and it is
+important to consider weaknesses in current forecasting methods in that
+context. In the beginning weather forecasts were also worse than simple
+baselines, but these forecasts have continually improved throughout the history
+of the field [@mcgill2012; @silver2012; @bauer2015]. One practice that lead to
+improvements in weather forecasts was that large numbers of forecasts were made
+publicly, allowing different approaches to be regularly assessed and refined
+[@mcgill2012; @silver2012]. To facilitate this kind of improvement, it is
+important for ecologists to start regularly making and evaluating real
+ecological forecasts, even if they perform poorly, and to make these forecasts
+openly available for assessment [@mcgill2012; @dietze2016]. These forecasts
+should include both short-term predictions, which can be assessed quickly, and
+mid- to long-term forecasts, which can help ecologists to assess long time-scale
+processes and determine how far into the future we can successfully forecast
+[@dietze2016; @tredennick2016]. Forecasts out to 2050 based on the models in
+this paper have been openly archived [@zenodo], so that we and others can assess
+how well they perform. We plan to evaluate these forecasts and report the
+results as each new year of BBS data becomes available, and make iterative
+improvements to the forecasting models in response to these assessments.
+
+Making successful ecological forecasts will be challenging. Ecological systems
+are complex, our fundamental theory is less refined than for simpler physical
+and chemical systems, and we currently lack the scale of data that often
+produces effective forecasts through machine learning. Despite this, we believe
+that progress can be made if we develop an active forecasting culture in ecology
+that builds and assesses forecasts in ways that will allow us to improve the
+effectiveness of ecological forecasts more rapidly [Box 1; @mcgill2012;
+@dietze2016]. This includes expanding the scope of the ecological and
+environmental data we work with, paying attention to uncertainty in both model
+building and forecast evaluation, and rigorously assessing forecasts using a
+combination of hindcasting, archived forecasts, and comparisons to simple
+baselines.
+
+## Acknowledgments
+
+This research was supported by the Gordon and Betty Moore Foundation's
+Data-Driven Discovery Initiative through Grant GBMF4563 to E.P. White. We thank
+the developers and providers of the data and software that made this research
+possible including: the PRISM Climate Group at Oregon State University, the
+staff at USGS and volunteer citizen scientists associated with the North
+American Breeding Bird Survey, NASA, the World Climate Research Programme's
+Working Group on Coupled Modelling and its working groups, the U.S. Department
+of Energy's Program for Climate Model Diagnosis and Intercomparison, and the
+Global Organization for Earth System Science Portals. A. C. Perry provided
+valuable comments that improved the clarity of this manuscript.
+
+## Box 1: Best practices for making and evaluating ecological forecasts
 
@@
 can learn more rapidly about the best approaches for making predictions for a
-given ecological quantity. This includes comparing process based and data-driven
-models and comparing the accuracy of forecasts to simple baselines to determine
-if the modeled forecasts are more accurate than the naive assumption that the
-world is static.
+given ecological quantity [@clark2001; @ward2014]. This includes comparing
+process-based [e.g., @kearney2009] and data-driven models [e.g., @ward2014], as
+well as comparing the accuracy of forecasts to simple baselines to determine if
+the modeled forecasts are more accurate than the naive assumption that the world
+is static [@jolliffe2003; @ye2015].
 
@@
 Forecasts describe how systems are expected to change through time. While some
-areas of ecological forecasting focus primarily on time-series data, others
-primarily focus on using spatial models and space-for time substitutions. Using
-ecological and environmental time-series data allows the consideration of actual
-dynamics from both a process [@treddenick2016a] and error structure perspective.
+areas of ecological forecasting focus primarily on time-series data [@ward2014], 
+others primarily focus on using spatial models and space-for-time substitutions
+[@blois2013]. Using ecological and environmental time-series data allows the
+consideration of actual dynamics from both a process and error structure
+perspective [@tredennick2016].
 
@@
 
-Understanding confidence in a forecast is just as important as understanding the
-average or expected outcome. Failing to account for uncertainty can result in
-overconfidence in highly uncertain outcomes leading to poor decision making and
-erosion of confidence in ecological forecasts. Models should explicitly include
-sources of uncertainty and how the propagate through the forecast where
-possible. Evaluations of forecasts should assess the accuracy of uncertainties
-as well as point estimates [@dietz2017].
+Understanding uncertainty in a forecast is just as important as understanding
+the average or expected outcome. Failing to account for uncertainty can result
+in overconfidence in uncertain outcomes leading to poor decision making and
+erosion of confidence in ecological forecasts [@clark2001]. Models should
+explicitly include sources of uncertainty and propagate them through the
+forecast where possible [@clark2001; @dietze2017]. Evaluations of forecasts
+should assess the accuracy of models' estimated uncertainties as well as their
+point estimates [@dietze2017].
 
@@
 include predictor variables that are expected to relate to the ecological
-quantity being forecast and dynamic time-series of predictors instead of
-long-term averages. Investing time in identifying and acquiring better predictor
+quantity being forecast. Time-series of predictors, instead of long-term
+averages, are also preferable to match the ecologial data (see #2).
+Investing time in identifying and acquiring better predictor
 variables may have at least as many benefits as using more sophisticated
-modeling techniques.
+modeling techniques [@kent2014].
 
-### 5. Assess how forecast accuracy changes with time-lag
+### 5. Address unknown or unmeasured predictors
 
-In general the accuracy of forecasts decreases with the length of time into the
-future being forecast [@petch2015]. This decay in accuracy and the potential for
-different rates of decay to result in different relative model performance at
-different lead times should be considered when evaluating forecasts and
-comparing models.
+Ecological systems are complex and many biotic and abiotic aspects of the
+environment are not regularly measured. As a result, some sites may deviate in
+consistent ways from model predictions. Unknown or unmeasured predictors can be
+incorporated in models using site-level random effects (potentially spatially
+autocorrelated) or by using latent variables that can identify unmeasured
+gradients [@harris2015].
 
-### 6. Include an observation model
+### 6. Assess how forecast accuracy changes with time-lag
 
-Ecological observations are influenced by both the underlying processes and how
-the system is sampled. When possible forecasts should model the factors
-influencing the observation of the data.
+In general, the accuracy of forecasts decreases with the length of time into the
+future being forecast [@petchey2015]. This decay in accuracy should be
+considered when evaluating forecasts. In addition to simple decreases in
+forecast accuracy the potential for different rates of decay to result in
+different relative model performance at different lead times should be
+considered.
 
-### 7. Validate using hindcasting
+### 7. Include an observation model
 
-To evaluate the expected accuracy and uncertainty of forecasts assess the
-performance of these forecasts within existing time-series data.
+Ecological observations are influenced by both the underlying biological
+processes (e.g. resource limitation) and how the system is sampled. When
+possible, forecasts should model the factors influencing the observation of the
+data [@yu2010; @hutchinson2011; @schurr2012].
 
-### 8. Publicly archive forecasts
+### 8. Validate using hindcasting
 
-To allow the future evaluation of the accuracy and uncertainty of forecasts the
-forecast values and/or models should be archived so that they can be assessed
-after new data is generated [@mcgill2012]. Enough information should be provided
-to allow an unambiguous assessment of the forecast performance.
+Evalutating a model's predictive performance across time is critical for
+understanding if it is useful for forecasting the future. Hindcasting uses a
+temporal out-of-sample validation approach to mimic how well a model would have
+performed had it been run in the past. For example, using occurance data from
+the early 20th century to model distributions which are validated with late 20th
+century occurances. Dense time series, such as yearly observations, are
+desirable to also evalulate the forecast horizon (see #6), but this is not a
+strict requirement.
 
-### 9. Make short-term and long-term predictions
+### 9. Publicly archive forecasts
 
-In cases where long-term predictions are the primary goal, short-term should
-also be made to accommodate the time-scales of planning and management decisions
-and to allow the accuracy of the forecasts to be quickly evaluated
-[@treddenick2016a].
+Forecast values and/or models should be archived so that they can be assessed
+after new data is generated [@mcgill2012; @silver2012; @dietze2016]. Enough
+information should be provided in the archive to allow unambiguous assessment of
+each forecast's performance [@tetlock2016].
 
-### 10. [[?]]
+### 10. Make both short-term and long-term predictions
 
-Sites' biotic and abiotic environments differ in more ways than the 
-few ones you happened to measure. As a result, some sites will 
-consistently have more species than a regression model (or sum of 
-regression models) would predict, while others willl consistently have 
-fewer. Depending on the spatial scale of these unmeasured variables, 
-independent site-level random effects could be sufficient for dealing 
-with this, or spatially-autocorrelated random effects might be needed.
+Even in cases where long-term predictions are the primary goal, short-term
+predictions should also be made to accommodate the time-scales of planning and
+management decisions and to allow the accuracy of the forecasts to be quickly
+evaluated [@dietze2016; @tredennick2016].
+
+## References

### 2017-11-20 manuscript/draft.Rmd
table formatting

@@
 author:
-- David J. Harris
-- Shawn D. Taylor
-- Ethan P. White
+- David J. Harris$^1$ (corresponding author)
+- Shawn D. Taylor$^2$
+- Ethan P. White$^1$
 bibliography: refs.bib
@@
 output:
-  pdf_document: default
+  pdf_document: 
+    keep_tex: true
   html_document: default
 ---
+
+$^1$ Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, FL, United States
+
+$^2$ School of Natural Resources and Environment, University of Florida Gainesville, FL, United States
 
@@
 
+\newpage
+
 ## Abstract
@@
 evaluating how well current models characterize natural systems. While the
-number of forecasts for biodiversity is increasing, there is little evidence
-showing that these forecasts work well at decadal time scales. Most biodiversity
-forecasts are not evaluated to determine how well they predict future diversity,
-fail to account for uncertainty, and do not use time-series data that captures
-the actual dynamics being studied. In this study, we used best practices and
-32-year time series to evaluate six different modeling approaches to predicting
-changes in North American breeding birds' species richness. Hindcasts for each
-method were evaluated annually for a decade at `r numbers$N_predict_sites` sites
-distributed throughout the continental United States. While each model could
-explain most of the variance in richness, none of them outperformed baseline
-models that predicted constant richness at each site. In particular, we found no
-evidence that current methods (such as species distribution models) can
-successfully turn spatial data into useful temporal predictions about
-biodiversity at decadal time-scales. To facilitate the rapid improvement of
-biodiversity forecasts, we emphasize the value of specific best practices at
-multiple stages of the forecasting pipeline, from the choice of predictor
-variables through the choice of approach for evaluating model performance over
-time.
+number of forecasts for biodiversity is increasing, there is little information
+available on how well these forecasts work. Most biodiversity forecasts are not
+evaluated to determine how well they predict future diversity, fail to account
+for uncertainty, and do not use time-series data that captures the actual
+dynamics being studied. We addressed these limitations by using best practices
+to explore our ability to forecast the species richness of breeding birds in
+North America. We used hindcasting to evaluate six different modeling approaches
+for predicting richness. Hindcasts for each method were evaluated annually for a
+decade at `r numbers$N_predict_sites` sites distributed throughout the
+continental United States. While each model could explain most of the variance
+in richness, none of them consistently outperformed a baseline model that
+predicted constant richness at each site. In particular, we found no evidence
+that current methods (such as species distribution models) can successfully turn
+spatial data into useful temporal predictions about biodiversity at decadal
+time-scales. The best practices implemented in this study directly influence the
+forecasts, the relative performance of different modeling approaches, and the
+conclusions about the current state of biodiversity forecasting . To facilitate
+the rapid improvement of biodiversity forecasts, we emphasize the value of
+specific best practices in making forecasts and evaluating forecasting methods.
 
@@
 We evaluated 6 types of forecasting models (Table 1) by dividing the 32 years
-of data into 22 years of training data and 10 years of data for 
-evaluating forecasts using hindcasting. We also made long term forecasts by using
-the full data set for training and making forecasts through the year 2050. For
-both time scales, we made forecasts using each model with and without
-correcting for observer effects, as described below.
+of data into 22 years of training data and 10 years of data for evaluating
+forecasts using hindcasting. Here we use definitions from meteorology, where a
+hindcast is generally any prediction for an event that has already happened,
+while forecasts are predictions for actual future events [@jolliffe2003]. We
+also made long term forecasts by using the full data set for training and
+making forecasts through the year 2050. For both time frames, we made forecasts
+using each model with and without correcting for observer effects, as described
+below.
 
@@
 Breeding Bird Survey (BBS) [@pardieck2017] using the Data Retriever Python
-package [@morris2013] and rdataretriever R package [@rdataretriever]. The BBS
-data was filtered to exclude all nocturnal, cepuscular, and aquatic species
-[since these species are not well sampled by BBS methods; @hurlbert2005], as
-well as unidentified species, and hybrids. All data from surveys that did not
-meet BBS quality criteria were also excluded.
+package [@morris2013] and rdataretriever R package [@rdataretriever]. BBS
+observations are three-minute point counts made at 50 fixed locations along a
+40km route. Here we denote each route as a site and summarize richness as the
+total species observed at all 50 locations in each surveyed year. Prior to
+summarizing the data was filtered to exclude all nocturnal, cepuscular, and
+aquatic species [since these species are not well sampled by BBS methods;
+@hurlbert2005], as well as unidentified species, and hybrids. All data from
+surveys that did not meet BBS quality criteria were also excluded.
 
@@
 
-table_data = read_csv("../figures/table.csv",col_types = "ccccc")
+table_data = read_csv("../figures/table.csv",col_types = "ccccc", na = "na")
 
@@
     booktabs = TRUE,
-    caption = table_caption
+    caption = table_caption,
+    align = c("l", "l", "c", "c", "c")
   ) %>%
@@
   group_rows("Single-site models", 1, 3) %>%
-  group_rows("Environmental models", 4, 6) %>% kable_styling(position = "center")
+  group_rows("Environmental models", 4, 6) %>% 
+  kable_styling(position = "center")
 ```
@@
 **Time series models.** We used Auto-ARIMA models [based on the `auto.arima`
-function in @forecast1] to represent an array of different time-series modeling
-approaches. These models can include an autoregressive component (as in the
-"naive" model, but with the possibility of longer-term dependencies in the
-underlying process), a moving average component (where the noise can have serial
-autocorrelation) and an integration/differencing component (so that the analysis
-could be performed on sequential differences of the raw data, accommodating more
-complex patterns including trends). The `auto.arima` function chooses whether to
-include each of these components (and how many terms to include for each one)
-using AICc [@forecast1]. Since there is no seasonal component to the BBS
-time-series, we did not include a season component in these models. Otherwise we
-used the default settings for this function [@forecast1].
+function in the package `forecast`; @forecast1] to represent an array of
+different time-series modeling approaches. These models can include an
+autoregressive component (as in the "naive" model, but with the possibility of
+longer-term dependencies in the underlying process), a moving average component
+(where the noise can have serial autocorrelation) and an
+integration/differencing component (so that the analysis could be performed on
+sequential differences of the raw data, accommodating more complex patterns
+including trends). The `auto.arima` function chooses whether to include each of
+these components (and how many terms to include for each one) using AICc
+[@forecast1]. Since there is no seasonal component to the BBS time-series, we
+did not include a season component in these models. Otherwise we used the
+default settings for this function [@forecast1].
 
@@
 for each of the species-level probabilities instead of just one, propagating the
-uncertainty forward. We obtained these estimates using random forests, a common
-approach in the species distribution modeling literature. Random forests are
-constructed by fitting hundreds of independent regression trees to
-randomly-perturbed versions of the data [@cutler2007; @caruana2008]. When
-correcting for observer effects, each of the 500 trees in our species-level
-random forests used a different Monte Carlo estimate of the observer effects as
-a predictor variable.
+uncertainty forward. We obtained these estimates using random forests
+[@randomForest], a common approach in the species distribution modeling
+literature. Random forests are constructed by fitting hundreds of independent
+regression trees to randomly-perturbed versions of the data [@cutler2007;
+@caruana2008]. When correcting for observer effects, each of the 500 trees in
+our species-level random forests used a different Monte Carlo estimate of the
+observer effects as a predictor variable.
 
@@
 (https://github.com/weecology/bbs-forecasting) and archived on Zenodo
-(https://doi.org/10.5281/zenodo.839581).
+[@zenodoCode].
 
@@
 
-Partitioning each model's squared error into consistent errors across sites and
-errors that fluctuated year-to-year shows that the majority of the error comes
-from errors in the site-level mean (Figure 6). The "average" model, which was
-based on the site-level mean, had the lowest error in this regard. In contrast,
-the three environmental models showed the largest biases at the site level; this
-makes sense, given that they could not explicitly distinguish among sites with
-similar climate, NDVI, and elevation.
+Partitioning each model's squared error shows that the majority of the residual
+error was attributed to errors in estimating site-level means, rather than
+errors in tracking year-to-year fluctuations (Figure 6). The "average" model,
+which was based entirely on site-level means, had the lowest error in this
+regard. In contrast, the three environmental models showed larger biases at the
+site level, though they still explained most of the variance in this component.
+This makes sense, given that they could not explicitly distinguish among sites
+with similar climate, NDVI, and elevation. Interestingly, the environmental
+models had higher squared error than the baselines did for tracking year-to-year
+fluctuations in richness as well.
 
@@
 fit (Figure 7). Improvements primarily resulted from a small number of forecasts
-where observer turnover caused a large shift in the reported richness values
-(Appendix [[Figure Number]]). The naive baseline was less sensitive to these
-shifts, because it largely ignored the richness values reported by observers
-that had retired by the end of the training period (Figure 1). The average
-model, which gave equal weight to observations from the whole training period,
-showed a larger decline in performance when not accounting for observer effects
--- especially in terms of coverage. The performance of the mistnet JSDM was
-notable here, because its prediction intervals retained good coverage even when
-not correcting for observer differences, which we attribute to the JSDM's
-ability to model this variation with its latent variables.
+where observer turnover caused a large shift in the reported richness values.
+The naive baseline was less sensitive to these shifts, because it largely
+ignored the richness values reported by observers that had retired by the end of
+the training period (Figure 1). The average model, which gave equal weight to
+observations from the whole training period, showed a larger decline in
+performance when not accounting for observer effects -- especially in terms of
+coverage. The performance of the mistnet JSDM was notable here, because its
+prediction intervals retained good coverage even when not correcting for
+observer differences, which we attribute to the JSDM's ability to model this
+variation with its latent variables.
 
@@
 When comparing different methods for forecasting our results demonstrate the
-importance of considering uncertainty [Box 1; @clark2001; @dietz2016]. Previous
+importance of considering uncertainty [Box 1; @clark2001; @dietze2016]. Previous
 comparisons between stacked SDMs and macroecological models reported that the
@@
 
-The accuracy for forecasts is expected to decline as forecasts are made further
-into the future, so it is important to evaluate forecasts at different lead
-times [Box 1; @petchey2015]. For deviance (the error metric that evaluates both
-models' point estimates and their uncertainty simultaneously), there is some
-indication that the JSDM and the nonlinear richness model converge towards the
-time-series and baseline models as the lead time of the forecast increases. This
-potential convergence is biologically plausible because at short lead times,
-environmental changes are relatively small and changes in the biota may lag
-behind changes in the environment. However, as the lead time of the forecast
-increases the amount of environmental change should increase and the systems
-will have more time to respond, potentially resulting in shifts towards improved
-relative performance of models incorporating environmental and ecological
-information [@blonder2017]. This is potentially important for interpreting our
-results because we only evaluated annual forecasts up to a decade into the
-future. In contrast, many ecological forecasts are temporally aggregated to 5-30
-year timesteps and have lead times of up to a century. These long-term forecasts
-are difficult to evaluate given the small number of century scale datasets, but
-are important for understanding changes in biodiversity at some of the lead
-times relevant for conservation and management. Two studies have assessed models
-of species richness at longer lead times [@algar2009; @distler2015], but the
-results were not compared to baseline or time-series models (in part due to data
-limitations) making them difficult to compare to our results. Given the
-potential changes with lead time of the models and processes are most relevant
-to forecasting, it is important that more long lead time evaluation studies are
-conducted and that the inferences drawn from short-term and long-term
-forecasting studies be compared. For true forecasts, short term forecasts will
-allow for relatively rapid validation of models by researchers without needing
-to wait decades for data to accumulate [@tredennick2016; @dietze2016]. However,
-since drivers of species richness differ at different temporal scales
-[@rosenzweig1995; @white2004; @white2007], short-term forecasts should be viewed
-as informing, not replacing, long-term forecasts [@tredennick2016].
+We have only evaluated annual forecasts up to a decade into the future, but
+forecasts are often made with a lead time of 50 years or more. These long-term
+forecasts are difficult to evaluate given the small number of century-scale
+datasets, but are important for understanding changes in biodiversity at some of
+the lead times relevant for conservation and management. Two studies have
+assessed models of species richness at longer lead times [@algar2009;
+@distler2015], but the results were not compared to baseline or time-series
+models (in part due to data limitations) making them difficult to compare to our
+results directly. Studies on shorter time scales, such as ours, provide one way
+to evaluate our forecasting methods without having to wait several decades to
+observe the effects of environmental change on biodiversity [@petchey2015;
+@tredennick2016; @dietze2016], but cannot fully replace longer-term evaluations
+[@tredennick2016]. In general, drivers of species richness can differ at
+different temporal scales [@rosenzweig1995; @white2004; @white2007;
+@blonder2017], so different methods may perform better for different lead times.
+In particular, we might expect environmental and ecological information to
+become more important at longer time scales, and thus for the performance of
+simple baseline forecasts to degrade faster than forecasts from SDMs and other
+similar models.  We did observe a small trend in this direction: deviance for
+the auto-ARIMA models and for the average baseline grew faster than for two of
+the environmental models (the JSDM and the macroecological model), although this
+growth was not statistically significant for the average baseline.
 
@@
 fits at longer lead times, it is also possible that they will continue to
-produce forecasts that are worse than baselines assuming that the systems are
+produce forecasts that are worse than baselines that assume the systems are
 static.  This would be expected to occur if richness in these systems is not
@@
 absolute accuracy of forecasts across different sites: when a past year's
-richness is already known, it is easy to estimate future richness.  The
-site-level stability of the BBS data also explains why SDMs and macroecological
-models perform relatively well at predicting future richness, despite failing to
+richness is already known, it is easy to estimate future richness. Ward et al.
+(2014) found similar patterns in time series of fisheries stocks, where
+relatively stable time series were best predicted by simple models and more
+complex models were only beneficial with dynamic time series. The site-level
+stability of the BBS data also explains why SDMs and macroecological models
+perform relatively well at predicting future richness, despite failing to
 capture changes in richness over time. However, this stability also makes it
@@
 system. These results suggest that single-site models should be actively
-considered for forecasts of richness and other stable aspects of
-biodiversity. Our results also suggest that future efforts to understand and
-forecast biodiversity should incorporate species composition, since lower-level
-processes are expected to be more dynamic [@ernest2001; @dornelas2014] and
-contain more useful information [@harris2015].
+considered for forecasts of richness and other stable aspects of biodiversity.
+Our results also suggest that future efforts to understand and forecast
+biodiversity should incorporate species composition, since lower-level processes
+are expected to be more dynamic [@ernest2001; @dornelas2014] and contain more
+useful information [@harris2015].
 
@@
 uncertainty and bias. Ultimately, ecological forecasts that use environmental
-data will therefore be more uncertain than our current hindcasting efforts. It
-is important to correctly incorporate this uncertainty in the predictor
-variables into forecasting models [@clark2001; @dietze2017]. Limitations in
-forecasting future environmental conditions, particularly at small scales, will
-present continued challenges for models incorporating environmental variables,
-and this may result in a continued advantage for simple single-site approaches.
+data will therefore be more uncertain than our current hindcasting efforts, and
+it is important to correctly incorporate this uncertainty into our models
+[@clark2001; @dietze2017]. Limitations in forecasting future environmental
+conditions---particularly at small scales---will present continued challenges
+for models incorporating environmental variables, and this may result in a
+continued advantage for simple single-site approaches.
 
@@
 effects resulted in changes in which models performed best, most notably
-reducing the relative performance of the naive model's point estimates. This
-suggests that, prior to accounting for observer effects, the naive model
-performed well because it was capable of accommodating rapid shifts in estimated
-richness introduced by changes in the observer. These kinds of rapid changes
-were difficult for the other single-site models to accommodate, and so the
-average performance of the ARIMA and average models improved once this source of
-observation error was addressed. This demonstrates that modeling observation
-error can be important for properly estimating and reducing uncertainty in
-forecasts and can also lead to changes in the best methods for forecasting [Box
-1]. Another key aspect of an ideal observation model is imperfect detection. In
-this study, we did not address differences in detection probability across
-species and sites [@boulinier1998] since there is no clear way to address this
-issue using NA Breeding Bird Survey data without making strong assumptions about
-the data [i.e., assuming there is no biological variation in stops along a
-route; @white2010], but this would be a valuable addition to future forecasting
-models.
+improving most models' point estimates relative to the naive baseline. This
+demonstrates that modeling observation error can be important for properly
+estimating and reducing uncertainty in forecasts and can also lead to changes in
+the best methods for forecasting [Box 1]. This suggests that, prior to
+accounting for observer effects, the naive model performed well largely because
+it was capable of accommodating rapid shifts in estimated richness introduced by
+changes in the observer. These kinds of rapid changes were difficult for the
+other single-site models to accommodate. Another key aspect of an ideal
+observation model is imperfect detection. In this study, we did not address
+differences in detection probability across species and sites [@boulinier1998]
+since there is no clear way to address this issue using North American Breeding
+Bird Survey data without making strong assumptions about the data [i.e.,
+assuming there is no biological variation in stops along a route; @white2010],
+but this would be a valuable addition to future forecasting models.
 
 The science of forecasting biodiversity remains in its infancy and it is
-important to consider weaknesses in current forecasting methods in that
-context. In the beginning weather forecasts were also worse than simple
-baselines, but these forecasts have continually improved throughout the history
-of the field [@mcgill2012; @silver2012; @bauer2015]. One practice that lead to
-improvements in weather forecasts was that large numbers of forecasts were made
-publicly, allowing different approaches to be regularly assessed and refined
-[@mcgill2012; @silver2012]. To facilitate this kind of improvement, it is
-important for ecologists to start regularly making and evaluating real
-ecological forecasts, even if they perform poorly, and to make these forecasts
-openly available for assessment [@mcgill2012; @dietze2016]. These forecasts
-should include both short-term predictions, which can be assessed quickly, and
-mid- to long-term forecasts, which can help ecologists to assess long time-scale
-processes and determine how far into the future we can successfully forecast
-[@dietze2016; @tredennick2016]. Forecasts out to 2050 based on the models in
-this paper have been openly archived [@zenodo], so that we and others can assess
-how well they perform. We plan to evaluate these forecasts and report the
-results as each new year of BBS data becomes available, and make iterative
-improvements to the forecasting models in response to these assessments.
+important to consider weaknesses in current forecasting methods in that context.
+In the beginning, weather forecasts were also worse than simple baselines, but
+these forecasts have continually improved throughout the history of the field
+[@mcgill2012; @silver2012; @bauer2015]. One practice that lead to improvements
+in weather forecasts was that large numbers of forecasts were made publicly,
+allowing different approaches to be regularly assessed and refined [@mcgill2012;
+@silver2012]. To facilitate this kind of improvement, it is important for
+ecologists to start regularly making and evaluating real ecological forecasts,
+even if they perform poorly, and to make these forecasts openly available for
+assessment [@mcgill2012; @dietze2016]. These forecasts should include both
+short-term predictions, which can be assessed quickly, and mid- to long-term
+forecasts, which can help ecologists to assess long time-scale processes and
+determine how far into the future we can successfully forecast [@dietze2016;
+@tredennick2016]. We have openly archived forecasts from all six models through
+the year 2050 [@zenodoPredict], so that we and others can assess how well they
+perform. We plan to evaluate these forecasts and report the results as each new
+year of BBS data becomes available, and make iterative improvements to the
+forecasting models in response to these assessments.
 

### 2017-11-20 manuscript/reviewer_reply.Rmd
law of total variance

## Editors Comments

*The reviewers and I agree that this manuscript is interesting, timely and well-written. I will add a few of my own observations here. First, and least important - normally submitted manuscripts do not hyphenate and do not right justify.*

*Next, in your figure 1, which forms the basis for comparison of model accuracy over time, I really expected you to choose data WITH a temporal pattern. After all, it looks to me like you have a time series in which species richness simply varies around a more or less stable average value, and the variation can be due to nothing (random), observer bias, and any number of species population trends (and combinations thereof). Thus, you have a situation of years of no trend, and use those data to predict no trend into the future? I would have thought you would choose data with a definite time trend that was already potentially correlated with, say, climate change, and then use hindcasting and forecasting to test whether the models accurately predict a trend into the "future"! To me, it is not surprising that models do not "...successfully turn spatial data into useful temporal predictions about biodiversity at decadal time-scales." But, I wonder if you chose data WITH a trend, you would have gotten other results.*

*Finally, your figure two seems to reverse panels A and B, neither of which is marked in the text.*

*I will leave the remainder of the comments to the two reviewers.*

## Reviewer 1: Eric Ward
#### Basic reporting

*I found the paper very clear and easy to follow, with a good intro/discussion that tied the current work into previous studies. All data shared - in addition to code. More minor comments in General Comments section.*
#### Experimental design

*This question and forecasting competition was well designed. I had minor comments to improve clarity with models. The analysis is all documented/replicable on the Github repo. More minor comments in General Comments section.*
#### Validity of the findings

*All data is robust, and analyses are sound. Conclusions and recommendations are well stated. The authors have really done a great job with transparency - this is a model of what all papers should strive for. See additional comments in General Comments section*

#### Comments for the author

#### Review of Harris et al. Forecasting biodiversity in breeding birds using best practices

*In general, I thought this paper was interesting and well written. These results add to the increasing literature supporting the idea that ecological forecasting in time/space is difficult. I think the recommendations/best practice bullet points described by the authors are essential for anyone interested in forecasting this kind of data. All of my points below can be considered minor issues, which I think will improve the paper’s clarity.*

#### Introduction:
*Line 41: I think this is generally more common in terrestrial ecology – in fisheries these SDM models with more of a theoretical background don’t seem to be used as much*

*Line 62: I really like the recommendations / clarity in Box 1, and think these will be accessible to most ecologists*

*Line 71-73: I’m re-reading this after commenting on data, and am now a little confused. The bird responses are annual snapshots during the breeding season? And I thought the environmental data you used was annual or quarterly, not monthly?* https://www.mbr-pwrc.usgs.gov/bbs/grass/genintro.htm

**BBS data are generally viewed as annual data, even though the observations happen on a single day. The presence of an individual species is primarily a function of suitable habitat and, for many species, timing of migration. Environmental variables with a shorter temporal scale could likely improve models, such as incorporating cool weather days into the observation model, and would be worth pursuing in a different study.**

#### Methods:

*Line 92: For those not familiar with the BBS protocol, maybe describe in a sentence. I could be mis-remembering, but isn’t each site made up of 50 stops, and presumably you aggregated richness across these stops?*

**This is correct, we've added clarification to the Methods.**

*Line 105: I assume all these species have been similarly ID’d over the course of the dataset – e.g. birds split into sub-species haven’t been included.*

**This is correct, see the section "Richness Data" in the Methods.**

*Line 107: For the environmental variables, do you have any that match up with the temporal resolution of the response data (monthly?). It seems like all the variables are annual / quarterly / etc.*

**See comment above on using shorter time scale environmental variables.**

*Line 124: I’m not as familiar with terrestrial projections, but are the CMIP5 projections relatively unbiased / precise when compared to recent data?*

*Line 138: Maybe also cite any of the interfaces you used, like RStan: http://mc-stan.org/users/citations/*

**RStan is cited in the last paragraph of the Methods.**

*Line 140: I’m curious why you didn’t consider spatial structure in the random effects here, using GMRF models or similar? One hand these models might be overkill if the habitat/environmental variables (which are spatially correlated) explain variability in richness, but the spatial correlation may pick up other sources of variability not accounted for by those variables.*

*Line 141-143: I think the paper thus far is pretty clear. I was hoping to see a few equations here – it helps me especially when components start to become hierarchical!*

*Line 163-164: again, I think equations could really help here to describe the baseline models. I think it’s also important to use these to clarify key details – for example, by training models on a site-by-site basis, you’re estimating separate parameters by site. For example with the uncorrelated noise model, you’re estimating unique variances by site right? Maybe say something like this, in addition to including equations.*

*Line 180 – Maybe also mention this is in the forecast package – as you do with gbm below?*

**We have added this.**

*Line 183: Because these defaults may change over time, it’d be good to specify things like the ranges of AR / MA order that you searched over, which criteria was used to compare models, etc. And whether or not you included a drift parameter?*

*Table 1 caption: without getting to more methods (below), I don’t understand the sentence ‘Environmental models were trained using all sites together, without information regarding which transects occurred at which site or during which year’.*

*Table 1: Just my personal preference, but I’d include the check marks and replace the ‘Xs’ with blank cells*

** We have made this change**

*Line 194-221: It seems for each approach you only considered the full models with all environmental covariates? I’m curious whether any of the covariates appeared to not improve explanatory power, and whether you considered models that dropped these uninformative predictors?*

*Line 216: Cite randomForest / caret/ party R package – or whatever else you used?*

**The randomForest package was cited in the final methods paragraph. We've cited it where it is first mentioned as well.**

*Line 242: Again, I think including the equation here would be useful*

**We have added an in-line equation and clarified the wording**

*Line 245: In our 2014 forecasting paper, we used MASE over RMSE because of rationale in Hyndmand and Koehler (2006). Using RMSE here is probably ok because you probably don’t have zeros/huge values/etc*

*Line 249: Can you elaborate on what you did with the deviance? By itself mean deviance is just a measure of average fit (I think most of the models are giving you this). Var(deviance) should incorporate precision – but I’ve only seen this extracted from Bayesian models*

*Line 266: I think you guys have done a great job putting together this public repository + made the analysis totally transparent. My one hang up with the repo was that it’d be awesome to include a short vignette with example data from this project that is more easily accessible to readers without having to install Jupyter/R packages/etc. For example – I was travelling when I started this review and couldn’t install one of the R packages on my government computer without making an appointment with our IT staff (so it took ~ a day). This isn’t your fault of course, but much harder for me to dig through the repo.*

*Line 269: See my above comments about spatially correlated random effects / GMRF models. One way to confirm these other approaches are overkill would be to make some maps of the estimated random effects, and fit some simple models to estimate variogram parameters – e.g. even gls() in R can do this*

*Line 276: Is one of the potential reasons that richness is stationary that presence/absence on the BBS varies less than other measures – maybe abundance, or other measures of diversity are more variable?*

*Line 294: Is the narrow predictive distribution for the stacked SDM also why this exhibits the strongest trend in decreasing coverge as a function of forecast length (Fig 5)?*

*Line 329 – maybe reword, you use forecast 3x in this sentence*

*Line 414: Agreed – doing this same kind of forecasting competition with compositional data, or other species diversity measures (Simpson’s) would be interesting*

*Line 461: Another thread with conservation implications is that back in the 1990s, it was pretty common for single species PVA models to be forecasted 100-200 years. This seems less common, in part because some predictions may have not performed well, and others are on a time horizon that can’t be evaluated. In Steve Beissinger’s PVA book he cites Goldwasser (2000) “Variability and measurement error in extinction risk analysis...” for this point (p 133).*

## Reviewer 2

#### Basic reporting

*see general comments*

#### Experimental design

*see general comments*

#### Validity of the findings

*see general comments*

#### Comments for the author

*In this paper, the authors use the incredibly rich data from the Breeding Bird Survey to examine how well various models perform at predicting bird richness at sites throughout North America. They train models on data from 1982-2003 and attempt to predict data from 2004-2013. They examine the performance of six different model types, two of which are null or baseline models.*

*I find it a bit surprising that the authors think that any of these richness models could predict a high level of accuracy for theses ten years. I think of richness estimations as a very coarse technique for looking at broad shifts in richness over relatively long time periods rather than these 5-species (~5-7%) shifts in richness on annual bases indicated in Figure 2. The authors don’t provide a lot of their raw data on sites other than as examples, but my impression is that they are pursuing an unrealistic level of accuracy and detail. That being said, given what the paper does, it’s okay. The models seem to be performed well. The authors give all of the keywords to indicate that they know what they are doing.*

*Here are some comments:*

*Birds are not the best system for doing any sort of distribution analyses. They are incredibly vagile, hard to detect, and the distributions of breeding birds are affected by many components besides environmental variables (particularly interspecific interactions). No assessment of detectability is included here other than accounting for observer identity. However, I do realize that BBS data are detailed and compelling data to use given what a large amount of detailed time-series it provides.*

*Abstract: The authors forecast the data and then test it using known data. Hindcast only refers to back-in-time predictions.*

**The definition of hindcasting varies by field. We use it here as it's used in meteorology, where any prediction of observations already available is generally a hindcast, and forecasts are only for predictions which are actually in the future. See Jolliffe & Stephenson 2003 (in manuscript) section 1.4.2 for a longer discussion. We've clarified this in the text.**

*25-29: rather than relay the type of information that the paper will provide, the authors should summarize their results and findings*

*57-58: this is not the definition of hindcasting. This is the definition of testing. Also, the authors exclude the many deeper-time hindcasting studies that have been performed on these types of models*

**See comment above on hindcasting.**

#### Methods:

*the authors perform their analyses on data that are collected from an incredibly dynamic time period. I wouldn’t expect species to have settled into a richness pattern given the incredibly disrupted nature of the last 32 years. Some strategies to try to handle these issues might be to train and predict at different temporal scales and to both forecast and hindcast the predictions. (e.g. train on 2004-2013 and try to predict 1982-1992 data + 1993-2003 data). If the authors were to make splits based on expected periods of dynamism, they could better tell whether the drivers that are included in the models are accurate predictors.*

*124-130: I didn’t see where future projections were performed/used in this paper. Why is this here?*

*Fig. 2: Panels not labeled. No key for colors. Somewhat confusing.*

*Fig. 6 caption: “compoonents”*

*466: only other mention of future forecasts… not sure why this is included in this paper.*

### 2017-12-07 manuscript/reviewer_reply.Rmd
more Ward commentes

@@
 First, and least important - normally submitted manuscripts do not hyphenate and do not right justify.*
+
+**Thank you. We have removed the right justification and hyphenation.**
 
@@
 *Finally, your figure two seems to reverse panels A and B, neither of which is marked in the text.*
+
+**We have switched the panels and labeled them. Thank you for noticing this error**.
 
@@
 
+**Thank you.**
+
 #### Introduction:
@@
 
+**This is a good point. We have added the word "terrestrial" to this line.**
+
 *Line 62: I really like the recommendations / clarity in Box 1, and think these will be accessible to most ecologists*
+
+**Thank you.**
 
@@
 
+**We have added the equations and Stan code as a supplemental file.**
+
 *Line 163-164: again, I think equations could really help here to describe the baseline models.
@@
 
-** We have made this change**
+**We have made this change**
 
@@
 I’m curious whether any of the covariates appeared to not improve explanatory power, and whether you considered models that dropped these uninformative predictors?*
+
+**We considered dropping uninformative predictors, but decided not to because of the complexity it would add to our analyses.
+In particular, it wasn't clear how to assess variable importance in a consistent way across three different model types and 500 Monte Carlo samples, and we did not want to bias the results by choosing a variable selection method that worked better for one method than another.**
 
@@
 Using RMSE here is probably ok because you probably don’t have zeros/huge values/etc*
+
+**We agree.**
 
@@
 
+>We agree that this is likely.
+Richness seems to be surprisingly stable in many systems, even when other measures change substantially, as we note on lines [[place we cite Dornelas et al.]] and [[[other place we cite Dornelas et al.]].
+
 *Line 294: Is the narrow predictive distribution for the stacked SDM also why this exhibits the strongest trend in decreasing coverge as a function of forecast length (Fig 5)?*
+
+>Yes, exactly. See lines [[XX]].
 
@@
 
+**Thank you for pointing out that we were unclear here.  The methods and purpose for the future predictions are now explained in more detail on lines [[XX]].**
+
 *Fig. 2: Panels not labeled. No key for colors. Somewhat confusing.*
+
+**Thank you for pointing this out.  As the Editor noted, we had inadvertently reversed the two panels. We have fixed this, and also labeled them as you suggested.**
 
@@
 
+**Thank you for pointing out this error. We have fixed it**
+
 *466: only other mention of future forecasts… not sure why this is included in this paper.*

### 2017-12-07 manuscript/reviewer_reply.Rmd

@@
 
-**We have switched the panels and labeled them. Thank you for noticing this error**.
+>We have switched the panels and labeled them. Thank you for noticing this error**.
 
@@
 
-**Thank you.**
+>Thank you.
 
@@
 
-**BBS data are generally viewed as annual data, even though the observations happen on a single day.
+>BBS data are generally viewed as annual data, even though the observations happen on a single day.
 The presence of an individual species is primarily a function of suitable habitat and, for many species, timing of migration.
-Environmental variables with a shorter temporal scale could likely improve models, such as incorporating cool weather days into the observation model, and would be worth pursuing in a different study.**
+Environmental variables with a shorter temporal scale could likely improve models, such as incorporating cool weather days into the observation model, and would be worth pursuing in a different study.
 
@@
 
-**This is correct, we've added clarification to the Methods.**
+>This is correct, we've added clarification to the Methods.
 
@@
 
-**This is correct, see the section "Richness Data" in the Methods.**
+>This is correct, see the section "Richness Data" in the Methods.
 
@@
 
-**See comment above on using shorter time scale environmental variables.**
+>See comment above on using shorter time scale environmental variables.
 
@@
 
+**The accuracy of CMIP5 models varies by the variable considered.
+Temperature is represented very well over North America.
+Precipitation is challenging, especially over mountainous terrain, but the CMIP5 ensemble represents precipitation across North America relatively well ("Evaluation of precipitation and temperature simulation performance of the CMIP3 and CMIP5 historical experiments", Koutroulis et al.
+2016).**
+
 *Line 138: Maybe also cite any of the interfaces you used, like RStan: http://mc-stan.org/users/citations/*
 
-**RStan is cited in the last paragraph of the Methods.**
+>RStan is cited in the last paragraph of the Methods.
 
@@
 
-**We have added this.**
+>We have added this.
 
@@
 
-**We have made this change**
+>We have made this change
 
@@
 
-**The randomForest package was cited in the final methods paragraph. We've cited it where it is first mentioned as well.**
+>The randomForest package was cited in the final methods paragraph. We've cited it where it is first mentioned as well.
 
@@
 
-**The definition of hindcasting varies by field.
+>The definition of hindcasting varies by field.
 We use it here as it's used in meteorology, where any prediction of observations already available is generally a hindcast, and forecasts are only for predictions which are actually in the future.
 See Jolliffe & Stephenson 2003 (in manuscript) section 1.4.2 for a longer discussion.
-We've clarified this in the text.**
+We've clarified this in the text.
 
@@
 
-**See comment above on hindcasting.**
+>See comment above on hindcasting.
 
@@
 
-**Thank you for pointing out that we were unclear here.  The methods and purpose for the future predictions are now explained in more detail on lines [[XX]].**
+>Thank you for pointing out that we were unclear here.  The methods and purpose for the future predictions are now explained in more detail on lines [[XX]].
 
@@
 
-**Thank you for pointing out this error. We have fixed it**
+>Thank you for pointing out this error. We have fixed it
 

### 2017-12-07 manuscript/reviewer_reply.Rmd

@@
 
-**Thank you. We have removed the right justification and hyphenation.**
+>Thank you. We have removed the right justification and hyphenation.
 
@@
 
->We have switched the panels and labeled them. Thank you for noticing this error**.
+>We have switched the panels and labeled them. Thank you for noticing this error.
 
@@
 
-**This is a good point. We have added the word "terrestrial" to this line.**
+>This is a good point. We have added the word "terrestrial" to this line.
 
@@
 
-**The accuracy of CMIP5 models varies by the variable considered.
+>The accuracy of CMIP5 models varies by the variable considered.
 Temperature is represented very well over North America.
 Precipitation is challenging, especially over mountainous terrain, but the CMIP5 ensemble represents precipitation across North America relatively well ("Evaluation of precipitation and temperature simulation performance of the CMIP3 and CMIP5 historical experiments", Koutroulis et al.
-2016).**
+2016).
 
@@
 
-**We have added the equations and Stan code as a supplemental file.**
+>We have added the equations and Stan code as a supplemental file.
 
@@
 
-**We considered dropping uninformative predictors, but decided not to because of the complexity it would add to our analyses.
-In particular, it wasn't clear how to assess variable importance in a consistent way across three different model types and 500 Monte Carlo samples, and we did not want to bias the results by choosing a variable selection method that worked better for one method than another.**
+>We considered dropping uninformative predictors, but decided not to because of the complexity it would add to our analyses.
+In particular, it wasn't clear how to assess variable importance in a consistent way across three different model types and 500 Monte Carlo samples, and we did not want to bias the results by choosing a variable selection method that worked better for one method than another.
 
@@
 
-**We have added an in-line equation and clarified the wording**
+>We have added an in-line equation and clarified the wording
 
@@
 
-**We agree.**
+>We agree.
 
@@
 
-**Thank you for pointing this out.  As the Editor noted, we had inadvertently reversed the two panels. We have fixed this, and also labeled them as you suggested.**
+>Thank you for pointing this out.  As the Editor noted, we had inadvertently reversed the two panels. We have fixed this, and also labeled them as you suggested.
 
